### 2019-08-13 main.tex
Adding current version of the paper from word file

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2018

% ready for submission
% \usepackage{neurips_2018}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2018}

% to compile a camera-ready version, add the [final] option, e.g.:
     \usepackage[final]{neurips_2018}

% to avoid loading the natbib package, add option nonatbib:
\usepackage[nonatbib]{neurips_2018}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\title{A data-driven study of cystic fibrosis treatment: identifying sub-types of physiotherapy style and physical activity using unsupervised clustering}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{%
  David S.~Hippocampus\thanks{Use footnote for providing further information
    about author (webpage, alternative address)---\emph{not} for acknowledging
    funding agencies.} \\
  Department of Computer Science\\
  Cranberry-Lemon University\\
  Pittsburgh, PA 15213 \\
  \texttt{hippo@cs.cranberry-lemon.edu} \\
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

\begin{document}

\maketitle

\begin{abstract}
  Children and young people with cystic fibrosis (CYPwCF) are prescribed physiotherapy including daily airway clearance techniques (ACT), and being physically active as part of their treatment. Little is known about how CYPwCF follow this advice in their everyday lives. Physiotherapy has evolved over the decades without this detailed data. Such data is now being collected as part of Project Fizzyo which is the largest CF physiotherapy trial in the UK to date. The data comes from novel pressure sensors attached to existing ACT devices which remotely measure breathing patterns during normal treatments, and Fitbit activity trackers which monitor physical activity patterns. In close collaboration with clinical experts, we quantify and characterise this unique ACT and activity data by creating custom features of the waveforms. We identify sub-types with unsupervised learning, with k-means clustering achieving the best performance. We find 4 clusters of ACT data which are sensitive enough to track interventions in ACT routine, 4 clusters of steps data, and 5 clusters of heartrate data, which are clinically meaningful sub-types.
\end{abstract}

\section{Problem overview}

Cystic Fibrosis (CF) is a chronic life limiting illness affecting approximately 1 in 2500 babies born globally. CF is a genetic disorder that mainly affects the lungs, where excessive thick sticky mucous leads to a cycle of infection, inflammation and lung damage leading to a deterioration in health. Despite recent progress CF remains progressive and incurable. In some cases, a lung-transplant is performed, but only half of the current CF population are expected to live to over 40 years old [ref].

To clear the lungs of excess mucous, daily physiotherapy airway clearance techniques (ACT) are prescribed. The physiotherapy exercises involve a series of forced exhalations into a handheld device  followed by a huff and cough to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly [ref to guideline?]. For children with CF, the physiotherapy is burdensome, with some reporting that they dislike their physiotherapy more than they dislike the disease itself [ref]. Physiotherapists also encourage children to have a physically active lifestyle [ref to guideline?]. Until recently adherence to physiotherapy recommendations was only assessed via questionnaire or self report which is known to be inaccurate. The evolution of CF physiotherapy of the decades has not been informed by this important evidence.

Bespoke pressure sensors were developed to attach to existing ACT devices and remotely monitor ACT breathing patterns. One way to reduce the treatment burden of ACT may be by using gamification. To investigate if gamification can change physiotherapy adherence or patterns, bespoke games were developed which are driven by pressure changes measured with the sensor [ref 1: verbal description, ref 2: video demonstrating the point https://customers.microsoft.com/en-us/story/731791-fizzyo]. Physical activity was also assessed using Fitbit activity trackers.

 145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify ACT and physical activity patterns (via wearing of a Fitbit activity tracker), and to study the effects of gamification. This is the largest CF physiotherapy trial in the UK [ref to meta-analysis of CF trials, or previously largest studies?], collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives. The broader goals of the study are to link physiotherapy behaviour patterns with clinical outcomes, in order to optimise and personalise treatment. In this paper we describe the challenge of i) quantifying ACT pressure sensor data and Fitbit activity data by creating custom features of the waveforms, and ii) discovering sub-types of ACTs and activity using clustering techniques. These daily sub-types can be tracked over time, as interventions such as gaming are introduced.   
\section{Methods}
\subsection{Trial design and analysis approach}
We used an interrupted time series design for the 18 month trial, so the effect on ACT patterns of introducing and removing an intervention can be assessed [ref to details of trial design]. Interventions include introducing feedback to participants on their ACT (number of breaths) on a tablet application, introducing games, and later removing these interventions. The data presented here is from the first 8x months of the trial. Participants are: 65x males and 80x females ranging in age from 6-16 years old, mainly from the ethnic group primarily affected by CF (Caucasian), from 3 London hospitals, and are voluntarily part of the study. All data is completely de-identified for analysis to all researchers.

Our analysis seeks to find patterns in physiotherapy data. Once these patterns are known, they can be linked with clinical outcomes to determine if certain behaviour patterns are preferable. We do not impose pre-conceived labels such as “good” or “bad” on waveforms, thus we leave open the possibility that what is currently prescribed may not be optimal. As such, the data is unlabelled and we use unsupervised learning approach to achieve this. Unsupervised clustering has been used to find subtypes in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].

A high-scale data analysis and exploration pipeline was built in R in a secure analysis environment.

\subsection{Collaboration, trust and interpretability}
Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included [ref https://www.nature.com/articles/s41598-018-29523-2]. To address the issues that arise from a lack of collaboration [ref to stuff on collaboration and interpretability] our methodology involves close collaboration between data scientists and CF experts, making every data decision jointly. We use a novel approach of involving at least one clinical expert in engineering “stand-ups” every day over the course of building the data analysis pipeline (7 months). We only consider features that are meaningful to clinicians, and we balance traditional statistical evaluation metrics, with what is interpretable to clinicians.

\subsection{Feature engineering}
\subsubsection{Breath data description}
An ACT treatment commonly involves around 10 sets of 10 exhalation breaths into the ACT device. Each set loosens mucous, so that after the set the participant can huff and cough to expel it. The treatment may be performed once or twice per day; participants do not have identical prescriptions. The ACT device is fitted with a wireless pressure sensor sampling at 10 Hz, recording a pressure value (cm H2O) and its time stamp. Participants each used 1-3 of 4 different ACT devices compatible with the sensor; the Acapella, Aerobia, PariPEP and AstraPEP.  
\subsubsection{Processing breath data}
Several pre-processing steps were performed on the data which were a function of the custom sensor used: including assembling data packets and removing duplicate values. The sensor also introduced significant non-linear baseline drift. Because of the asymmetrical nature of the exhalation peaks and sparsity of the signal, we used a sparsity-based de-noising approach with an asymmetry factor, called BEADS [ref https://www.sciencedirect.com/science/article/abs/pii/S0169743914002032?via 3Dihub], which is designed for use with chromatogram waveforms which have a similar form. 
% figure 1
In total, we recorded 5x,000 sensor readings. This includes data from every time the ACT sensor was turned on, including for testing and demonstration. Once we removed readings shorter than 15 seconds, treatments with no breath exhalations, and situations with sensor misuse/malfunction, we were left with 2x,000 ACT treatments for analysis.

\subsubsection{Featurising breath data}
% figure 2
The defining characteristics of ACT treatments were summarised in a range of features (variables): breath count, breath duration, breath amplitude, and treatment duration. In order to derive these features, our breath identification algorithm recognises breaths composed of multiple peaks, and having breath shapes which are characteristic of the 3 different ACT devices used in the study. In addition to mean breath amplitude, we use amplitude standard deviation as a feature, in order to capture inconsistency or disorderliness in physiotherapy. We also identified how many discernible sets were apparent in the treatment. These features are on a per-treatment level.

The aforementioned features capture the characteristics of existing treatments. However it is common for prescribed treatments to be skipped, and days with no treatments is an important signal itself. Several features were derived which capture missing data: treatment adherence to date, skipped days since last treatment, rolling adherence over 7 days. These features are on a per-day level, as participants are prescribed 1 or 2 treatments per day.

\subsubsection{Featurising physical activity data}
The Fitbit activity tracker (Fitbit AltaHR) provides each participant’s step count per minute. While real-time heartbeat data would allow for a detailed analysis of heartrate waveforms, the Fitbit only provides average heartrate, with a variable sampling frequency ranging from 3 to 12 values per minute. To achieve consistent frequency and comparable values, the heartrate values are averaged for each minute.

Mindful of these data limitations, we extract high-level patterns from the minute averaged waveforms at a day level. The heartrate and steps datasets are featurised using summary statistics per day using counts, frequency and interval values such as number of minutes in moderate to vigorous intensity physical activities (MVPA) (heartrate above 120 bpm) per day, step cadence (number of steps/minutes of wear) at day level, and number of hours with more than 500 steps at day level.

% figure 3

\subsubsection{Missing activity data}

To account for missing data during times that the Fitbit is not worn, the wear percent of the Fitbit is calculated from the heartrate values. This measure is used to select only days with a substantial portion of data for the day (80% of wear time from 6 am to 6 pm) for analysis. The wear time is also used to normalise some features and to obtain step cadence for minutes where heartrate data is present.

\subsubsection{Clustering to find sub-types}

All features were verified by clinicians as being meaningful or potentially meaningful. We perform a correlation analysis on all features and removed features which were highly correlated. We use the Hopkins statistic [ref] to measure clustering tendency. Features with a Gaussian distribution are normalised to have mean =0 and standard deviation 1. In order to visualise the results, we use dimensionality reduction. We use the first two principle components from principal component analysis [ref], Uniform Manifold Approximation and Projection (UMAP), and t-Distributed Stochastic Neighbor Embedding (T-SNE). We identify outliers using xxx methods and remove less than 2\%, with these cases being recognised as device malfunctions or unusual human behaviour.

We employ the following clustering methods: k-means, Density-based spatial clustering of applications with noise (DBSCAN) and hierarchical clustering.

\section{Results}
\subsection{Clustering breath waveforms}
\subsubsection{Feature selection for clustering}
Some features were excluded from clustering, but provide interesting physiotherapy insights.  
The number of sets in a treatment were expected to be very influential in clustering, as a regular pattern of sets shows close adherence to physiotherapy advice. However we found that less than 20xx\% of treatments had discernible sets, so this feature could not be used

% table 1

\subsubsection{ACT clustering results}
The results of the four techniques used for clustering are show in table x. K-means was found to perform best because we liked the way..… K-means is a simple

% table 2
% table 3
% figure 4

\subsubsection{ACT clustering validation}

We have checked the clustering stability on different data partitions.  Random sampling of the data (80\% of data) and using k-means clustering provided very similar results comparing to clustering on the full dataset:  shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index  (2200 v 1770 respectively) are comparable. 
% figure
We also have done validation of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters. Figure xx shows the same patient trajectory for the Clustering A and Clustering B. Clustering B uses breath amplitude features (mean and standard dev) as well as breathRollWeekAdherenceScore in addition to features that Clustering A uses. Notable that  once computer gaming is introduced to ACT in both cases A and B the same patient moves from cluster with “expected” breath counts per ACT to cluster characterized by much bigger breath counts. This validates that there is a change in patient’s behaviour that we’ve captured. Additional sdBreathAmplitude feature make results more explainable from clinical point of view as original cluster for the patient in Clustering B was characterized by centroid with the biggest sdBreathAmplitude value. 
% figure

\subsection{Clustering physical activity waveforms}

Clustering of physical activity data was used to extract daily patterns from the steps and heartrate waveforms that can be utilised to individually characterise the physical activity levels and patterns of individual patients. The different clusters should extract some more complex characteristics than the ones which could be obtained by simply combining some of the extracted features.

\subsubsection{Steps clusters}

The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, the mean of total step count per hour, the step cadence for the periods in MVPA and the number of hours in the day with 0<steps count < 500. Features were selected as a result of feature selection, correlation and outlier analysis. Four clusters were found by Kmeans with the following characteristics:

\begin{itemize}
\item XS - this cluster contains waveforms with very little steps activity or much missing step data 
\item S - moderate step activity and consistent cadence throughout the day with very few periods of high step activity 
\item M - consistent moderate to high step rate across the day with period of sustained high step activity  
\item L - High step rate throughout the day with longer periods of sustained high step activity
\end{itemize}
%figure
%table

subsubsection{Heartrate clusters}

The number of minutes with heartrate higher than 120, coefficient of variation for heartrate values at a day level, number of time heartrate goes above 100 and 120 and the median average hourly heartrate were used as features for the heartrate clusters. Five clusters were found by Kmeans with the following chacteristics: 
\begin{itemize}
\item S - Small heartrate values and consistently below 120 
\item M - Consistent variations around 100 with very few spikes in heartrate values 
\item M1 - Waveform with high variability and periods with spikes above 120 and fluctuations above the 120 mark 
\item M2 - Consistent variations around 100 with spikes in values above 120 
\item L - Large values around and above 120 
\end{itemize}
%figure
%table

\subsubsection{Fitbit cluster validation }

The steps and heartrate clusters were validated with the clinical experts. Individual waveforms were manually inspected together with the cluster boundaries to ensure consistency across cluster interpretations and the actual cluster characteristics. Additionally, other features that were not included in the clustering analysis were used to validate the clusters and their interpretations. Certain expected conditions were checked against the cluster labels and plotted using the PCA graph such as Figure Xxx below.

\section{Discussion and conclusion}

We have developed a technique of feature engineering and clustering analysis to bring insights into novel data about

We have demonstrated feature engineering and clustering analysis on novel data from a cystic fibrosis trial.

We have analysed the first 7 months of data in the 18 month...

\bibliographystyle{abbrv}
% include your .bib file
\bibliography{fizzyo}

\end{document}

### 2019-08-15 draft.tex
Updates from Overleaf

--- 2019-08-13 main.tex
@@
 \usepackage{microtype}      % microtypography
+\usepackage{graphicx}
 
@@
 
-\author{%
-  David S.~Hippocampus\thanks{Use footnote for providing further information
-    about author (webpage, alternative address)---\emph{not} for acknowledging
-    funding agencies.} \\
-  Department of Computer Science\\
-  Cranberry-Lemon University\\
-  Pittsburgh, PA 15213 \\
-  \texttt{hippo@cs.cranberry-lemon.edu} \\
+%\author{%
+%  David S.~Hippocampus\thanks{Use footnote for providing further information
+%    about author (webpage, alternative address)---\emph{not} for acknowledging
+%    funding agencies.} \\
+%    I think we need to keep anonymous for our submission\\
+%  Department of Computer Science\\
+%  Cranberry-Lemon University\\
+%  Pittsburgh, PA 15213 \\
+%  \texttt{hippo@cs.cranberry-lemon.edu} \\
   % examples of more authors
@@
   % \texttt{email} \\
-}
+%}
 
@@
 Physiotherapy has evolved over the decades without this detailed data.
-Such data is now being collected as part of Project Fizzyo which is the largest CF physiotherapy trial in the UK to date.
+Such data is now being collected as part of Project [anonymous] which is the largest CF physiotherapy trial in the UK to date.
 The data comes from novel pressure sensors attached to existing ACT devices which remotely measure breathing patterns during normal treatments, and Fitbit activity trackers which monitor physical activity patterns.
@@
 Despite recent progress CF remains progressive and incurable.
-In some cases, a lung-transplant is performed, but only half of the current CF population are expected to live to over 40 years old [ref].
+In some cases, a lung-transplant is performed, but only half of the current CF population are expected to live to over 40 years old [*******UCL****** ref to this or similar].
 
 To clear the lungs of excess mucous, daily physiotherapy airway clearance techniques (ACT) are prescribed.
-The physiotherapy exercises involve a series of forced exhalations into a handheld device  followed by a huff and cough to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly [ref to guideline?].
-For children with CF, the physiotherapy is burdensome, with some reporting that they dislike their physiotherapy more than they dislike the disease itself [ref].
-Physiotherapists also encourage children to have a physically active lifestyle [ref to guideline?].
-Until recently adherence to physiotherapy recommendations was only assessed via questionnaire or self report which is known to be inaccurate.
+The physiotherapy exercises involve a series of forced exhalations into a handheld device  followed by a huff and cough to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly [*******UCL****** ref to guideline?].
+For children with CF, the physiotherapy is burdensome, with some reporting that they dislike their physiotherapy more than they dislike the disease itself [*******UCL****** ref?
+This was from a survey that Jackie mentioned].
+Physiotherapists also encourage children to have a physically active lifestyle [*******UCL****** ref to why physical activity is good].
+Until recently adherence to physiotherapy recommendations was only assessed via questionnaire or self report which is known to be inaccurate [*******UCL****** ref].
 The evolution of CF physiotherapy of the decades has not been informed by this important evidence.
@@
 One way to reduce the treatment burden of ACT may be by using gamification.
-To investigate if gamification can change physiotherapy adherence or patterns, bespoke games were developed which are driven by pressure changes measured with the sensor [ref 1: verbal description, ref 2: video demonstrating the point https://customers.microsoft.com/en-us/story/731791-fizzyo].
+To investigate if gamification can change physiotherapy adherence or patterns, bespoke games were developed which are driven by pressure changes measured with the sensor [ref 1: verbal description of games, ref 2: video demonstrating the point https://customers.microsoft.com/en-us/story/731791-fizzyo, removing any identifyin names in the bibliography].
 Physical activity was also assessed using Fitbit activity trackers.
@@
  145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify ACT and physical activity patterns (via wearing of a Fitbit activity tracker), and to study the effects of gamification.
-This is the largest CF physiotherapy trial in the UK [ref to meta-analysis of CF trials, or previously largest studies?], collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives.
+This is the largest CF physiotherapy trial in the UK [*******UCL****** ref to meta-analysis of CF trials, or previously largest studies, or any previous data-driven studies?], collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives.
 The broader goals of the study are to link physiotherapy behaviour patterns with clinical outcomes, in order to optimise and personalise treatment.
@@
 \subsection{Trial design and analysis approach}
-We used an interrupted time series design for the 18 month trial, so the effect on ACT patterns of introducing and removing an intervention can be assessed [ref to details of trial design].
+We used an interrupted time series design for the 18 month trial, so the effect on ACT patterns of introducing and removing an intervention can be assessed [[*******UCL****** ref to details of trial design,is this sufficient?
+NREC:18-LO-1038 ].
 Interventions include introducing feedback to participants on their ACT (number of breaths) on a tablet application, introducing games, and later removing these interventions.
 The data presented here is from the first 8x months of the trial.
-Participants are: 65x males and 80x females ranging in age from 6-16 years old, mainly from the ethnic group primarily affected by CF (Caucasian), from 3 London hospitals, and are voluntarily part of the study.
-All data is completely de-identified for analysis to all researchers.
+Participants are: [*******UCL****** exact numbers] 65x males and 80x females ranging in age from 6-16 years old, from 3 London hospitals, and are voluntarily part of the study.
+All data is completely de-identified for analysis to researchers.
 
@@
 The sensor also introduced significant non-linear baseline drift.
-Because of the asymmetrical nature of the exhalation peaks and sparsity of the signal, we used a sparsity-based de-noising approach with an asymmetry factor, called BEADS [ref https://www.sciencedirect.com/science/article/abs/pii/S0169743914002032?via 3Dihub], which is designed for use with chromatogram waveforms which have a similar form.
+Because of the asymmetrical nature of the exhalation peaks and sparsity of the signal, we used a sparsity-based de-noising approach with an asymmetry factor, called BEADS \cite{Ning2014}, which is designed for use with chromatogram waveforms which have a similar form.
 % figure 1
@@
 To account for missing data during times that the Fitbit is not worn, the wear percent of the Fitbit is calculated from the heartrate values.
-This measure is used to select only days with a substantial portion of data for the day (80% of wear time from 6 am to 6 pm) for analysis.
+This measure is used to select only days with a substantial portion of data for the day (80\% of wear time from 6 am to 6 pm) for analysis.
 The wear time is also used to normalise some features and to obtain step cadence for minutes where heartrate data is present.
@@
 Clustering of physical activity data was used to extract daily patterns from the steps and heartrate waveforms that can be utilised to individually characterise the physical activity levels and patterns of individual patients.
-The different clusters should extract some more complex characteristics than the ones which could be obtained by simply combining some of the extracted features.
+The different clusters should extract more complex characteristics than the ones which could be obtained by simply combining some of the extracted features.
 
@@
 
-The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, the mean of total step count per hour, the step cadence for the periods in MVPA and the number of hours in the day with 0<steps count < 500.
+The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, the mean of total step count per hour, the step cadence for the periods in MVPA and the number of hours in the day with 0 < steps count < 500.
 Features were selected as a result of feature selection, correlation and outlier analysis.
@@
 
-We have developed a technique of feature engineering and clustering analysis to bring insights into novel data about
+Project Fizzyo involves collecting novel data about how CF physiotherapy advice, ACTs and physical activity, is adhered to.
+In close collaboration with CF experts, we develop methods to interrogate this data.
+We derive clinically meaningful features from ACT and activity (heartrate and steps) waveforms.
+We use unsupervised clustering to find sub-types of this data.
 
-We have demonstrated feature engineering and clustering analysis on novel data from a cystic fibrosis trial.
+Project Fizzyo is an ongoing study, with participation lasting 18 months.
+7 months into the trial we do not draw any conclusions, as we do not have an equal number of days of data from patients.
+Instead, we present a framework for analysis, which will empower clinicians to interpret the impact of interventions such as gamification, and what type of physiotherapy leads to better clinical outcomes.
 
-We have analysed the first 7 months of data in the 18 month...
+\section{Usage of tables and figures}
+\subsection{Figures}
+
+\begin{figure}
+  \centering
+  \caption{Sample figure caption.}
+  \includegraphics[]{sample_fig}
+\end{figure}
+
+\subsection{Tables}
+\begin{table}
+  \caption{Sample table title}
+  \label{sample_table}
+  \centering
+  \begin{tabular}{lll}
+    \toprule
+    \cmidrule(r){1-2}
+    Col1 & Col2 & Col3 \\
+    \midrule
+    a & 0 & z\\
+    b & 1 & y \\
+    c & 2 & x \\
+    \bottomrule
+  \end{tabular}
+\end{table}
+
+Here is how you can reference a table \ref{sample_table}
 

### 2019-08-15 final.tex
Updates from Overleaf

--- 2019-08-13 main.tex
@@
 \begin{abstract}
-  Children and young people with cystic fibrosis (CYPwCF) are prescribed physiotherapy including daily airway clearance techniques (ACT), and being physically active as part of their treatment.
-Little is known about how CYPwCF follow this advice in their everyday lives.
-Physiotherapy has evolved over the decades without this detailed data.
-Such data is now being collected as part of Project Fizzyo which is the largest CF physiotherapy trial in the UK to date.
-The data comes from novel pressure sensors attached to existing ACT devices which remotely measure breathing patterns during normal treatments, and Fitbit activity trackers which monitor physical activity patterns.
-In close collaboration with clinical experts, we quantify and characterise this unique ACT and activity data by creating custom features of the waveforms.
-We identify sub-types with unsupervised learning, with k-means clustering achieving the best performance.
-We find 4 clusters of ACT data which are sensitive enough to track interventions in ACT routine, 4 clusters of steps data, and 5 clusters of heartrate data, which are clinically meaningful sub-types.
+  Abstract here
 \end{abstract}
 
-\section{Problem overview}
-
-Cystic Fibrosis (CF) is a chronic life limiting illness affecting approximately 1 in 2500 babies born globally.
-CF is a genetic disorder that mainly affects the lungs, where excessive thick sticky mucous leads to a cycle of infection, inflammation and lung damage leading to a deterioration in health.
-Despite recent progress CF remains progressive and incurable.
-In some cases, a lung-transplant is performed, but only half of the current CF population are expected to live to over 40 years old [ref].
-
-To clear the lungs of excess mucous, daily physiotherapy airway clearance techniques (ACT) are prescribed.
-The physiotherapy exercises involve a series of forced exhalations into a handheld device  followed by a huff and cough to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly [ref to guideline?].
-For children with CF, the physiotherapy is burdensome, with some reporting that they dislike their physiotherapy more than they dislike the disease itself [ref].
-Physiotherapists also encourage children to have a physically active lifestyle [ref to guideline?].
-Until recently adherence to physiotherapy recommendations was only assessed via questionnaire or self report which is known to be inaccurate.
-The evolution of CF physiotherapy of the decades has not been informed by this important evidence.
-
-Bespoke pressure sensors were developed to attach to existing ACT devices and remotely monitor ACT breathing patterns.
-One way to reduce the treatment burden of ACT may be by using gamification.
-To investigate if gamification can change physiotherapy adherence or patterns, bespoke games were developed which are driven by pressure changes measured with the sensor [ref 1: verbal description, ref 2: video demonstrating the point https://customers.microsoft.com/en-us/story/731791-fizzyo].
-Physical activity was also assessed using Fitbit activity trackers.
-
- 145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify ACT and physical activity patterns (via wearing of a Fitbit activity tracker), and to study the effects of gamification.
-This is the largest CF physiotherapy trial in the UK [ref to meta-analysis of CF trials, or previously largest studies?], collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives.
-The broader goals of the study are to link physiotherapy behaviour patterns with clinical outcomes, in order to optimise and personalise treatment.
-In this paper we describe the challenge of i) quantifying ACT pressure sensor data and Fitbit activity data by creating custom features of the waveforms, and ii) discovering sub-types of ACTs and activity using clustering techniques.
-These daily sub-types can be tracked over time, as interventions such as gaming are introduced.
-\section{Methods}
-\subsection{Trial design and analysis approach}
-We used an interrupted time series design for the 18 month trial, so the effect on ACT patterns of introducing and removing an intervention can be assessed [ref to details of trial design].
-Interventions include introducing feedback to participants on their ACT (number of breaths) on a tablet application, introducing games, and later removing these interventions.
-The data presented here is from the first 8x months of the trial.
-Participants are: 65x males and 80x females ranging in age from 6-16 years old, mainly from the ethnic group primarily affected by CF (Caucasian), from 3 London hospitals, and are voluntarily part of the study.
-All data is completely de-identified for analysis to all researchers.
-
-Our analysis seeks to find patterns in physiotherapy data.
-Once these patterns are known, they can be linked with clinical outcomes to determine if certain behaviour patterns are preferable.
-We do not impose pre-conceived labels such as “good” or “bad” on waveforms, thus we leave open the possibility that what is currently prescribed may not be optimal.
-As such, the data is unlabelled and we use unsupervised learning approach to achieve this.
-Unsupervised clustering has been used to find subtypes in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].
-
-A high-scale data analysis and exploration pipeline was built in R in a secure analysis environment.
-
-\subsection{Collaboration, trust and interpretability}
-Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included [ref https://www.nature.com/articles/s41598-018-29523-2].
-To address the issues that arise from a lack of collaboration [ref to stuff on collaboration and interpretability] our methodology involves close collaboration between data scientists and CF experts, making every data decision jointly.
-We use a novel approach of involving at least one clinical expert in engineering “stand-ups” every day over the course of building the data analysis pipeline (7 months).
-We only consider features that are meaningful to clinicians, and we balance traditional statistical evaluation metrics, with what is interpretable to clinicians.
-
-\subsection{Feature engineering}
-\subsubsection{Breath data description}
-An ACT treatment commonly involves around 10 sets of 10 exhalation breaths into the ACT device.
-Each set loosens mucous, so that after the set the participant can huff and cough to expel it.
-The treatment may be performed once or twice per day; participants do not have identical prescriptions.
-The ACT device is fitted with a wireless pressure sensor sampling at 10 Hz, recording a pressure value (cm H2O) and its time stamp.
-Participants each used 1-3 of 4 different ACT devices compatible with the sensor; the Acapella, Aerobia, PariPEP and AstraPEP.
-\subsubsection{Processing breath data}
-Several pre-processing steps were performed on the data which were a function of the custom sensor used: including assembling data packets and removing duplicate values.
-The sensor also introduced significant non-linear baseline drift.
-Because of the asymmetrical nature of the exhalation peaks and sparsity of the signal, we used a sparsity-based de-noising approach with an asymmetry factor, called BEADS [ref https://www.sciencedirect.com/science/article/abs/pii/S0169743914002032?via 3Dihub], which is designed for use with chromatogram waveforms which have a similar form.
-% figure 1
-In total, we recorded 5x,000 sensor readings.
-This includes data from every time the ACT sensor was turned on, including for testing and demonstration.
-Once we removed readings shorter than 15 seconds, treatments with no breath exhalations, and situations with sensor misuse/malfunction, we were left with 2x,000 ACT treatments for analysis.
-
-\subsubsection{Featurising breath data}
-% figure 2
-The defining characteristics of ACT treatments were summarised in a range of features (variables): breath count, breath duration, breath amplitude, and treatment duration.
-In order to derive these features, our breath identification algorithm recognises breaths composed of multiple peaks, and having breath shapes which are characteristic of the 3 different ACT devices used in the study.
-In addition to mean breath amplitude, we use amplitude standard deviation as a feature, in order to capture inconsistency or disorderliness in physiotherapy.
-We also identified how many discernible sets were apparent in the treatment.
-These features are on a per-treatment level.
-
-The aforementioned features capture the characteristics of existing treatments.
-However it is common for prescribed treatments to be skipped, and days with no treatments is an important signal itself.
-Several features were derived which capture missing data: treatment adherence to date, skipped days since last treatment, rolling adherence over 7 days.
-These features are on a per-day level, as participants are prescribed 1 or 2 treatments per day.
-
-\subsubsection{Featurising physical activity data}
-The Fitbit activity tracker (Fitbit AltaHR) provides each participant’s step count per minute.
-While real-time heartbeat data would allow for a detailed analysis of heartrate waveforms, the Fitbit only provides average heartrate, with a variable sampling frequency ranging from 3 to 12 values per minute.
-To achieve consistent frequency and comparable values, the heartrate values are averaged for each minute.
-
-Mindful of these data limitations, we extract high-level patterns from the minute averaged waveforms at a day level.
-The heartrate and steps datasets are featurised using summary statistics per day using counts, frequency and interval values such as number of minutes in moderate to vigorous intensity physical activities (MVPA) (heartrate above 120 bpm) per day, step cadence (number of steps/minutes of wear) at day level, and number of hours with more than 500 steps at day level.
-
-% figure 3
-
-\subsubsection{Missing activity data}
-
-To account for missing data during times that the Fitbit is not worn, the wear percent of the Fitbit is calculated from the heartrate values.
-This measure is used to select only days with a substantial portion of data for the day (80% of wear time from 6 am to 6 pm) for analysis.
-The wear time is also used to normalise some features and to obtain step cadence for minutes where heartrate data is present.
-
-\subsubsection{Clustering to find sub-types}
-
-All features were verified by clinicians as being meaningful or potentially meaningful.
-We perform a correlation analysis on all features and removed features which were highly correlated.
-We use the Hopkins statistic [ref] to measure clustering tendency.
-Features with a Gaussian distribution are normalised to have mean =0 and standard deviation 1.
-In order to visualise the results, we use dimensionality reduction.
-We use the first two principle components from principal component analysis [ref], Uniform Manifold Approximation and Projection (UMAP), and t-Distributed Stochastic Neighbor Embedding (T-SNE).
-We identify outliers using xxx methods and remove less than 2\%, with these cases being recognised as device malfunctions or unusual human behaviour.
-
-We employ the following clustering methods: k-means, Density-based spatial clustering of applications with noise (DBSCAN) and hierarchical clustering.
-
-\section{Results}
-\subsection{Clustering breath waveforms}
-\subsubsection{Feature selection for clustering}
-Some features were excluded from clustering, but provide interesting physiotherapy insights.  
-The number of sets in a treatment were expected to be very influential in clustering, as a regular pattern of sets shows close adherence to physiotherapy advice.
-However we found that less than 20xx\% of treatments had discernible sets, so this feature could not be used
-
-% table 1
-
-\subsubsection{ACT clustering results}
-The results of the four techniques used for clustering are show in table x. K-means was found to perform best because we liked the way..… K-means is a simple
-
-% table 2
-% table 3
-% figure 4
-
-\subsubsection{ACT clustering validation}
-
-We have checked the clustering stability on different data partitions.
-Random sampling of the data (80\% of data) and using k-means clustering provided very similar results comparing to clustering on the full dataset:  shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index  (2200 v 1770 respectively) are comparable.
-% figure
-We also have done validation of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters.
-Figure xx shows the same patient trajectory for the Clustering A and Clustering B.
-Clustering B uses breath amplitude features (mean and standard dev) as well as breathRollWeekAdherenceScore in addition to features that Clustering A uses.
-Notable that  once computer gaming is introduced to ACT in both cases A and B the same patient moves from cluster with “expected” breath counts per ACT to cluster characterized by much bigger breath counts.
-This validates that there is a change in patient’s behaviour that we’ve captured.
-Additional sdBreathAmplitude feature make results more explainable from clinical point of view as original cluster for the patient in Clustering B was characterized by centroid with the biggest sdBreathAmplitude value.
-% figure
-
-\subsection{Clustering physical activity waveforms}
-
-Clustering of physical activity data was used to extract daily patterns from the steps and heartrate waveforms that can be utilised to individually characterise the physical activity levels and patterns of individual patients.
-The different clusters should extract some more complex characteristics than the ones which could be obtained by simply combining some of the extracted features.
-
-\subsubsection{Steps clusters}
-
-The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, the mean of total step count per hour, the step cadence for the periods in MVPA and the number of hours in the day with 0<steps count < 500.
-Features were selected as a result of feature selection, correlation and outlier analysis.
-Four clusters were found by Kmeans with the following characteristics:
-
-\begin{itemize}
-\item XS - this cluster contains waveforms with very little steps activity or much missing step data 
-\item S - moderate step activity and consistent cadence throughout the day with very few periods of high step activity 
-\item M - consistent moderate to high step rate across the day with period of sustained high step activity  
-\item L - High step rate throughout the day with longer periods of sustained high step activity
-\end{itemize}
-%figure
-%table
-
-subsubsection{Heartrate clusters}
-
-The number of minutes with heartrate higher than 120, coefficient of variation for heartrate values at a day level, number of time heartrate goes above 100 and 120 and the median average hourly heartrate were used as features for the heartrate clusters.
-Five clusters were found by Kmeans with the following chacteristics:
-\begin{itemize}
-\item S - Small heartrate values and consistently below 120 
-\item M - Consistent variations around 100 with very few spikes in heartrate values 
-\item M1 - Waveform with high variability and periods with spikes above 120 and fluctuations above the 120 mark 
-\item M2 - Consistent variations around 100 with spikes in values above 120 
-\item L - Large values around and above 120 
-\end{itemize}
-%figure
-%table
-
-\subsubsection{Fitbit cluster validation }
-
-The steps and heartrate clusters were validated with the clinical experts.
-Individual waveforms were manually inspected together with the cluster boundaries to ensure consistency across cluster interpretations and the actual cluster characteristics.
-Additionally, other features that were not included in the clustering analysis were used to validate the clusters and their interpretations.
-Certain expected conditions were checked against the cluster labels and plotted using the PCA graph such as Figure Xxx below.
-
-\section{Discussion and conclusion}
-
-We have developed a technique of feature engineering and clustering analysis to bring insights into novel data about
-
-We have demonstrated feature engineering and clustering analysis on novel data from a cystic fibrosis trial.
-
-We have analysed the first 7 months of data in the 18 month...
-
-\bibliographystyle{abbrv}
-% include your .bib file
-\bibliography{fizzyo}
-
 \end{document}

### 2019-08-16 draft.tex
Adding Fitbit info to appendix

@@
 
+\appendix
+\section{Appendix : Table of physical activity data features}
+
+\begin{table}[h]
+  \caption{Physical activity features}
+  \label{fitbit_features}
+  \centering
+  \begin{tabular}{ |l|l|l| }
+    \toprule
+    \cmidrule(r){1-2}
+    Feature Name & Feature description & Waveform extracted from \\
+    \midrule
+    userif & 1099 & 0.95 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\section{Appendix : Cluster metrics for physical activity data}
+
+The tables below show the metrics for the clusters identified for the physical activity data.
+
+\textbf{Steps clusters}
+
+Calinski-harabasz index for the steps clusters is 6021.14 .
+
+\begin{table}[h]
+  \caption{Steps clusters Metrics}
+  \label{steps_metrics}
+  \centering
+  \begin{tabular}{ |l|c|c|}
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster name & Cluster Size & Average Silhouette width \\
+    \midrule
+    XS & 1099 & 0.95 \\
+    S & 3942 & 0.34 \\
+    M & 2737 & 0.27 \\
+    L & 1982 & 0.24 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\textbf{Heartrate clusters}
+
+Calinski-harabasz index for the heartrate clusters is 5090.37 .
+
+\begin{table}[h]
+  \caption{Heartrate clusters Metrics}
+  \label{hr_metrics}
+  \centering
+  \begin{tabular}{ |l|c|c|}
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster name & Cluster Size & Average Silhouette width \\
+    \midrule
+    S & 1246 & 0.28 \\
+    SM & 1640 & 0.2 \\
+    M1 & 2308 & 0.21 \\
+    M2 & 2584 & 0.26 \\
+    L & 1982 & 0.33 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
 \end{document}

### 2019-08-31 draft.tex
ACT Features

@@
 \begin{abstract}
-  Children and young people with cystic fibrosis (CYPwCF) are prescribed physiotherapy including daily airway clearance techniques (ACT), and being physically active as part of their treatment.
+  Children and young people with cystic fibrosis (CYPwCF) are prescribed physiotherapy including daily airway clearance techniques (ACT), and physical activity as part of their treatment.
 Little is known about how CYPwCF follow this advice in their everyday lives.
@@
 The data comes from novel pressure sensors attached to existing ACT devices which remotely measure breathing patterns during normal treatments, and Fitbit activity trackers which monitor physical activity patterns.
-In close collaboration with clinical experts, we quantify and characterise this unique ACT and activity data by creating custom features of the waveforms.
-We identify sub-types with unsupervised learning, with k-means clustering achieving the best performance.
-We find 4 clusters of ACT data which are sensitive enough to track interventions in ACT routine, 4 clusters of steps data, and 5 clusters of heartrate data, which are clinically meaningful sub-types.
+In close collaboration with clinical experts, we quantify and characterise this unique ACT and activity data by creating custom features from the waveforms.
+Using unsupervised learning techniques, We identify meaningful sub-types, with k-means clustering achieving the best performance.
+We find 4 clusters of ACT data which are sensitive enough to track interventions in ACT routine.
+As pertaining to physical activity we identify 4 clusters of steps data and 5 clusters of heartrate data.
 \end{abstract}
@@
 
-Cystic Fibrosis (CF) is a chronic life limiting illness affecting approximately 1 in 2500 babies born globally.
-CF is a genetic disorder that mainly affects the lungs, where excessive thick sticky mucous leads to a cycle of infection, inflammation and lung damage leading to a deterioration in health.
+Cystic Fibrosis (CF) is a chronic life limiting illness affecting approximately 1 in 2500 babies born in the UK.
+CF is a systemic genetic disorder that mainly affects the respiratory and digestive systems.
+In the lungs excessive thick sticky mucus causes to a cycle of infection, inflammation and lung damage leading to a deterioration in health.
 Despite recent progress CF remains progressive and incurable.
@@
 
-To clear the lungs of excess mucous, daily physiotherapy airway clearance techniques (ACT) are prescribed.
-The physiotherapy exercises involve a series of forced exhalations into a handheld device  followed by a huff and cough to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly [*******UCL****** ref to guideline?].
+To promote clearance of mucus from the lungs, daily physiotherapy including airway clearance techniques (ACT) are prescribed.
+The physiotherapy exercises involve a series of  exhalations against resistance into a handheld device  followed by a huff and cough to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly \cite{Villanuevaj4574} [*******UCL****** ref to guideline?].
 For children with CF, the physiotherapy is burdensome, with some reporting that they dislike their physiotherapy more than they dislike the disease itself [*******UCL****** ref?
 This was from a survey that Jackie mentioned].
-Physiotherapists also encourage children to have a physically active lifestyle [*******UCL****** ref to why physical activity is good].
-Until recently adherence to physiotherapy recommendations was only assessed via questionnaire or self report which is known to be inaccurate [*******UCL****** ref].
-The evolution of CF physiotherapy of the decades has not been informed by this important evidence.
+Physiotherapists also encourage children to have a physically active lifestyle, this has been shown to reduce the rate of decline in lung function [*******UCL****** ref to why physical activity is good].
+Until recently adherence to physiotherapy recommendations was only assessed via questionnaire or self report which is known to be inaccurate with more objective methods not widely available [*******UCL****** ref].
+The evolution of CF physiotherapy has not been informed by this important evidence.
 
@@
 One way to reduce the treatment burden of ACT may be by using gamification.
-To investigate if gamification can change physiotherapy adherence or patterns, bespoke games were developed which are driven by pressure changes measured with the sensor [ref 1: verbal description of games, ref 2: video demonstrating the point https://customers.microsoft.com/en-us/story/731791-fizzyo, removing any identifyin names in the bibliography].
-Physical activity was also assessed using Fitbit activity trackers.
+To investigate if gamification can change ACT adherence or patterns, bespoke games were developed which are driven by pressure changes measured with the sensor [ref 1: verbal description of games, ref 2: video demonstrating the point https://customers.microsoft.com/en-us/story/731791-fizzyo, removing any identifyin names in the bibliography].
+Physical activity patterns were also measured using Fitbit activity trackers.
 
- 145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify ACT and physical activity patterns (via wearing of a Fitbit activity tracker), and to study the effects of gamification.
+ 145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify routine ACT (using bespoke pressure sensors) and physical activity patterns (via wearing of a Fitbit activity tracker), and to study the effects of gamification.
 This is the largest CF physiotherapy trial in the UK [*******UCL****** ref to meta-analysis of CF trials, or previously largest studies, or any previous data-driven studies?], collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives.
@@
 As such, the data is unlabelled and we use unsupervised learning approach to achieve this.
-Unsupervised clustering has been used to find subtypes in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].
+Unsupervised clustering has been used to find sub-types in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].
 
@@
 \subsection{Collaboration, trust and interpretability}
-Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included [ref https://www.nature.com/articles/s41598-018-29523-2].
-To address the issues that arise from a lack of collaboration [ref to stuff on collaboration and interpretability] our methodology involves close collaboration between data scientists and CF experts, making every data decision jointly.
+Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included \cite{Alaa2018}.
+To address the issues that arise from a lack of collaboration \cite{Vayena2018}, \cite{Challen231}, \cite{Char2018}, \cite{Ahmad2018} our methodology involves close collaboration between data scientists and CF experts, making every data decision jointly.
 We use a novel approach of involving at least one clinical expert in engineering “stand-ups” every day over the course of building the data analysis pipeline (7 months).
@@
     \cmidrule(r){1-2}
-    Feature Name & Feature description & Waveform extracted from \\
+    \textbf{Feature Name} & \textbf{Feature description} & \textbf{Waveform extracted from} \\
     \midrule
-    userif & 1099 & 0.95 \\
+    Active Hours & Number of hours with a step count > 500 & Steps waveform \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\appendix
+\section{Appendix : Table of ACT treatment features}
+
+\begin{table}[h]
+  \caption{ACT features}
+  \label{act_features}
+  \centering
+  \begin{tabular}{ |l|l| }
+    \toprule
+    \cmidrule(r){1-2}
+    \textbf{Feature Name} & \textbf{Feature description} \\
+    \midrule
+    Breath Count & Number of breaths the person has made though an ACT  \\ 
+     ~ & device during a treatment. A breath is counted if its maximum amplitude is \\ ~ & above the threshold of 8 mm H2O.\\
+     \midrule
+     Mean Breath Amplitude & Average breath amplitude (in mm H2O) during an ACT treatment. \\
+     Minimum Breath Amplitude & Similarly minimum, maximum and standard deviation of breath amplitude. \\
+     Maximum Breath Amplitude & ~ \\
+     Standard Deviation of Breath Amplitude & ~\\
+     \midrule
+     Mean Breath Duration & Average duration (in seconds) of breath during an ACT treatment.\\ 
+     Minimum Breath Duration &  Similarly minimum, maximum and standard deviation of \\ 
+     Maximum Break Duration & breath duration in a treatment. \\
+     Standard Deviation of Breath Duration & ~\\
+     \midrule
+     Break Count & Number of breaks (between breaths) in a treatment. \\
+     \midrule
+     Mean Break Duration & Average duration (in seconds) of breaks between breaths.\\ 
+     Minimum Break Duration & Similarly minimum, maximum and standard deviation of \\ 
+     Maximum Break Duration & breaks duration in a treatment. \\
+     Standard Deviation of Break Duration & ~\\
+     \midrule
+     Set Count & Number of sets in a treatment. Prescribed treatment may have up to 10 \\
+     ~ & sets. Set is a group of breaths separated by huffing and coughing activity.\\
+     \midrule
+     Pressure Readings Count & Number of pressure reading registered by ACT device during a treatment. \\
+     ~ & Used to filter out "test" device sessions that are not part of a treatment.\\
+     \midrule
+     Pressures Readings Completeness Score & Data completeness score for the day \\
+     ~ & If a treatment has less than 150 pressure readings it's completeness \\
+     ~ & score is 0. \\
+     \midrule
+     Treatment Duration & Treatment duration (in minutes).\\
+     \midrule
+     Treatment Daily Adherence Score & Number of completed treatments out of the prescribed ones (for the \\
+     ~ & given day). \\
+     \midrule
+     Breath Count Daily Adherence Score & Number of registered breaths across all treatments for the given day \\
+     ~ & out of the prescribed amount of breaths. \\ 
+     \midrule
+     Treatment Rolling Weekly Adherence Score & Number of completed treatments during the last 7 days (for a given date) \\
+     ~ & out of the prescribed number of treatments for this period. \\
+     \midrule
+     Breath Count Rolling Weekly Adherence Score & Number of registered breaths across all treatments during the last 7 days\\
+     ~ & (for a given date) out of the prescribed amount of breaths for this period.\ 
+     \midrule
+     Treatment Weekly Adherence Score & Number of completed treatments during the last calendar week (for a given date) \\
+     ~ & out of the prescribed number of treatments for this period. \\
+     \midrule
+     Breath Count Weekly Adherence Score & Number of registered breaths across all treatments during the last calendar week \\
+     ~ & (for a given date) out of the prescribed amount of breaths for this period.\\
+     \midrule
+     Cumulative Adherence Score & Cumulative adherence score to date. \\
+     ~ & How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount.\\
+     \midrule
+     Breathing Time Percentage & Time in a treatment spent on an exhalation, vs not.\\
+     \midrule
+     Skipped Days Count & Skipped days since the last treatment. \\
      \bottomrule

### 2019-09-02 draft.tex
Fix non-rendering issue, add ACT Clustering res image.

@@
 % figure 4
+Figure ~\ref{fig:act_res}  shows the four ACT clusters overlaid with a single participant (black), showing that the participant’s cluster membership changes over time.
+Clustering results are visualised in 2-dimensions using PCA (left) and UMAP (right).
+The time points are before ACT feedback (number of registered breaths) is introduced (top), after ACT feedback is introduced (middle), and once computer gaming is introduced to ACT routine (bottom).
+Each point represents one treatment of one participant.
+\begin{figure}[!htb]
+  \centering
+  \caption{ACT clusters}
+  \includegraphics[]{fig_ACT_clust}
+  \label{fig:act_res}
+\end{figure}
 
@@
 
-\begin{figure}
-  \centering
-  \caption{Sample figure caption.}
-  \includegraphics[]{sample_fig}
-\end{figure}
-
 \subsection{Tables}
@@
      Maximum Breath Amplitude & ~ \\
-     Standard Deviation of Breath Amplitude & ~\\
+     Standard Deviation of Breath Amplitude & ~ \\
      \midrule
-     Mean Breath Duration & Average duration (in seconds) of breath during an ACT treatment.\\ 
+     Mean Breath Duration & Average duration (in seconds) of breath during an ACT treatment. \\ 
      Minimum Breath Duration &  Similarly minimum, maximum and standard deviation of \\ 
      Maximum Break Duration & breath duration in a treatment. \\
-     Standard Deviation of Breath Duration & ~\\
+     Standard Deviation of Breath Duration & ~ \\
      \midrule
@@
      \midrule
-     Mean Break Duration & Average duration (in seconds) of breaks between breaths.\\ 
+     Mean Break Duration & Average duration (in seconds) of breaks between breaths. \\ 
      Minimum Break Duration & Similarly minimum, maximum and standard deviation of \\ 
      Maximum Break Duration & breaks duration in a treatment. \\
-     Standard Deviation of Break Duration & ~\\
+     Standard Deviation of Break Duration & ~ \\
      \midrule
      Set Count & Number of sets in a treatment. Prescribed treatment may have up to 10 \\
-     ~ & sets. Set is a group of breaths separated by huffing and coughing activity.\\
+     ~ & sets. Set is a group of breaths separated by huffing and coughing activity. \\
      \midrule
      Pressure Readings Count & Number of pressure reading registered by ACT device during a treatment. \\
-     ~ & Used to filter out "test" device sessions that are not part of a treatment.\\
+     ~ & Used to filter out "test" device sessions that are not part of a treatment. \\
      \midrule
@@
      \midrule
-     Treatment Duration & Treatment duration (in minutes).\\
+     Treatment Duration & Treatment duration (in minutes). \\
      \midrule
@@
      \midrule
-     Breath Count Rolling Weekly Adherence Score & Number of registered breaths across all treatments during the last 7 days\\
-     ~ & (for a given date) out of the prescribed amount of breaths for this period.\ 
+     Breath Count Rolling Weekly Adherence Score & Number of registered breaths across all treatments during the last 7 days \\
+     ~ & (for a given date) out of the prescribed amount of breaths for this period. \\ 
      \midrule
@@
      Cumulative Adherence Score & Cumulative adherence score to date. \\
-     ~ & How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount.\\
+     ~ & How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount. \\
      \midrule

### 2019-09-05 draft.tex
pushing all work for up to Sept 4th.

--- 2019-08-16 draft.tex
@@
 \usepackage[nonatbib]{neurips_2018}
-
 \usepackage[utf8]{inputenc} % allow utf-8 input
@@
 \usepackage{graphicx}
+\usepackage{float}
+\usepackage{longtable}
+\usepackage{array}
+\newcolumntype{n}{>{\arraybackslash}m{5.3cm}}
+\newcolumntype{d}{>{\arraybackslash}m{8cm}}
 
-\title{A data-driven study of cystic fibrosis treatment: identifying sub-types of physiotherapy style and physical activity using unsupervised clustering}
+\title{A data-driven study of physiotherapy for cystic fibrosis: identifying sub-types of airway clearance and physical activity patterns using unsupervised clustering}
 
@@
 Little is known about how CYPwCF follow this advice in their everyday lives.
-Physiotherapy has evolved over the decades without this detailed data.
-Such data is now being collected as part of Project [anonymous] which is the largest CF physiotherapy trial in the UK to date.
-The data comes from novel pressure sensors attached to existing ACT devices which remotely measure breathing patterns during normal treatments, and Fitbit activity trackers which monitor physical activity patterns.
-In close collaboration with clinical experts, we quantify and characterise this unique ACT and activity data by creating custom features of the waveforms.
-We identify sub-types with unsupervised learning, with k-means clustering achieving the best performance.
-We find 4 clusters of ACT data which are sensitive enough to track interventions in ACT routine, 4 clusters of steps data, and 5 clusters of heartrate data, which are clinically meaningful sub-types.
+Physiotherapy has evolved over the decades without the clarity of evidence that would accrue from detailed data.
+Such data are now being collected as part of Project [Anonymous] which is the largest CF physiotherapy trial in the UK to date.
+The data come firstly from novel pressure sensors attached to existing ACT devices which measure breathing patterns during normal treatments, and secondly from Fitbit activity trackers which monitor physical activity patterns.
+In close collaboration with clinical experts, we quantified and characterised this unique ACT and activity data by creating custom features from the waveforms.
+Using unsupervised learning techniques, We identified meaningful sub-types, with k-means clustering achieving the best performance.
+We found 4 clusters of ACT data which were sensitive enough to track interventions in ACT routines.
+Within physical activity data we identified 4 clusters related to daily footsteps  and 5 clusters related to heart rate data.
 \end{abstract}
@@
 
-Cystic Fibrosis (CF) is a chronic life limiting illness affecting approximately 1 in 2500 babies born globally.
-CF is a genetic disorder that mainly affects the lungs, where excessive thick sticky mucous leads to a cycle of infection, inflammation and lung damage leading to a deterioration in health.
-Despite recent progress CF remains progressive and incurable.
-In some cases, a lung-transplant is performed, but only half of the current CF population are expected to live to over 40 years old [*******UCL****** ref to this or similar].
+Cystic Fibrosis (CF) is the most common life-limiting inherited disorder in the UK, affecting approximately 1 in 2500 babies born.
+It is a systemic genetic disorder that mainly affects the respiratory and digestive systems.
+In the lungs excessive thick sticky mucus can cause cycles of infection, inflammation and lung damage leading to a deterioration in health.
+Despite improvements in care, CF remains progressive and incurable.
+In some cases, lung-transplantation can prolong life, but only half of the current CF population are expected to live to over 40 years old \cite{Keogh2018}.
 
-To clear the lungs of excess mucous, daily physiotherapy airway clearance techniques (ACT) are prescribed.
-The physiotherapy exercises involve a series of forced exhalations into a handheld device  followed by a huff and cough to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly [*******UCL****** ref to guideline?].
-For children with CF, the physiotherapy is burdensome, with some reporting that they dislike their physiotherapy more than they dislike the disease itself [*******UCL****** ref?
-This was from a survey that Jackie mentioned].
-Physiotherapists also encourage children to have a physically active lifestyle [*******UCL****** ref to why physical activity is good].
-Until recently adherence to physiotherapy recommendations was only assessed via questionnaire or self report which is known to be inaccurate [*******UCL****** ref].
-The evolution of CF physiotherapy of the decades has not been informed by this important evidence.
+To promote clearance of mucus from the lungs, daily physiotherapy is prescribed \cite{Daniels2017}.
+Airway clearance techniques (ACT) may include a series of  exhalations against resistance into hand-held ACT devices, followed by forced expiration techniques ('huffing') and coughing to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly \cite{Villanuevaj4574}.
+ACT is burdensome, with some CYPwCF reporting that ACT is the most burdensome part of having the disease \cite{JamesLindAlliance}.
+Physiotherapists also encourage children to have a physically active lifestyle, which has been shown to reduce the rate of decline in lung function \cite{Williams2014}.
+Adherence to physiotherapy has traditionally been assessed via questionnaire or self report, both of which are known to be inaccurate.
+More objective measures of adherence  have not been widely available, and the evolution of CF physiotherapy has not been informed by this important evidence \cite{Modi2006}.
 
-Bespoke pressure sensors were developed to attach to existing ACT devices and remotely monitor ACT breathing patterns.
-One way to reduce the treatment burden of ACT may be by using gamification.
-To investigate if gamification can change physiotherapy adherence or patterns, bespoke games were developed which are driven by pressure changes measured with the sensor [ref 1: verbal description of games, ref 2: video demonstrating the point https://customers.microsoft.com/en-us/story/731791-fizzyo, removing any identifyin names in the bibliography].
-Physical activity was also assessed using Fitbit activity trackers.
+Bespoke pressure sensors were developed to attach to existing ACT devices in order to facilitate remote monitoring of ACT undertaken routinely at home.
+In addition, bespoke games were developed which could be driven by pressure changes from breathing through the sensor.
+This was in order to investigate if gamification might reduce the treatment burden of ACT or modify ACT adherence or patterns \cite{Jackson2019} \cite{[Anonymised]2019} .
+Physical activity patterns were also measured using Fitbit activity trackers, which generated granular detail on footsteps and heart rate changes during the day.
 
- 145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify ACT and physical activity patterns (via wearing of a Fitbit activity tracker), and to study the effects of gamification.
+ 145 CYPwCF participated in the study [NREC:18-LO-1038] to quantify routine ACTs (using bespoke pressure sensors) and physical activity patterns (from wearing a Fitbit activity tracker), and to study the effects of gamification.
 This is the largest CF physiotherapy trial in the UK [*******UCL****** ref to meta-analysis of CF trials, or previously largest studies, or any previous data-driven studies?], collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives.
-The broader goals of the study are to link physiotherapy behaviour patterns with clinical outcomes, in order to optimise and personalise treatment.
-In this paper we describe the challenge of i) quantifying ACT pressure sensor data and Fitbit activity data by creating custom features of the waveforms, and ii) discovering sub-types of ACTs and activity using clustering techniques.
+The broader goals of the study are to link physiotherapy adherence patterns with clinical outcomes, in order to optimise and personalise treatment.
+In this paper we describe the challenge of i) quantifying ACT pressure sensor data and Fitbit activity data by creating custom features of the waveforms, and ii) discovering sub-types of ACT and physical activity patterns using clustering techniques.
 These daily sub-types can be tracked over time, as interventions such as gaming are introduced.
+
 \section{Methods}
+
 \subsection{Trial design and analysis approach}
-We used an interrupted time series design for the 18 month trial, so the effect on ACT patterns of introducing and removing an intervention can be assessed [[*******UCL****** ref to details of trial design,is this sufficient?
+
+The 18 month trial used an interrupted time series design, so that the effect of introducing and removing interventions such as gamification could be assessed [[*******UCL****** ref to details of trial design,is this sufficient?
 NREC:18-LO-1038 ].
-Interventions include introducing feedback to participants on their ACT (number of breaths) on a tablet application, introducing games, and later removing these interventions.
+Interventions included: introducing feedback to participants on their ACTs (number of breaths per treatment) via a computer tablet application, introducing games, and later removing these interventions.
 The data presented here is from the first 8x months of the trial.
-Participants are: [*******UCL****** exact numbers] 65x males and 80x females ranging in age from 6-16 years old, from 3 London hospitals, and are voluntarily part of the study.
-All data is completely de-identified for analysis to researchers.
+Participants were: [*******UCL****** exact numbers] 65x males and 80x females ranging in age from 6-16 years old, from 3 London hospitals, all of whom consented to participate in the study.
+All data were completely de-identified for analysis.
 
@@
 As such, the data is unlabelled and we use unsupervised learning approach to achieve this.
-Unsupervised clustering has been used to find subtypes in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].
+Unsupervised clustering is becoming an established technique for finding sub-types in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].
 
@@
 \subsection{Collaboration, trust and interpretability}
-Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included [ref https://www.nature.com/articles/s41598-018-29523-2].
-To address the issues that arise from a lack of collaboration [ref to stuff on collaboration and interpretability] our methodology involves close collaboration between data scientists and CF experts, making every data decision jointly.
-We use a novel approach of involving at least one clinical expert in engineering “stand-ups” every day over the course of building the data analysis pipeline (7 months).
-We only consider features that are meaningful to clinicians, and we balance traditional statistical evaluation metrics, with what is interpretable to clinicians.
+
+Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included \cite{Alaa2018}.
+To address the issues that arise from a lack of collaboration \cite{Vayena2018}, \cite{Challen231}, \cite{Char2018}, \cite{Ahmad2018} our methodology involved close collaboration between data scientists and CF experts, making every data decision jointly.
+We used a novel approach of involving at least one clinical expert in engineering “stand-ups” every day over the course of building the data analysis pipeline (7 months).
+We only considered features that were meaningful to clinicians, and we balanced traditional statistical evaluation metrics, with what was interpretable to clinicians.
 
 \subsection{Feature engineering}
+
 \subsubsection{Breath data description}
-An ACT treatment commonly involves around 10 sets of 10 exhalation breaths into the ACT device.
-Each set loosens mucous, so that after the set the participant can huff and cough to expel it.
-The treatment may be performed once or twice per day; participants do not have identical prescriptions.
-The ACT device is fitted with a wireless pressure sensor sampling at 10 Hz, recording a pressure value (cm H2O) and its time stamp.
-Participants each used 1-3 of 4 different ACT devices compatible with the sensor; the Acapella, Aerobia, PariPEP and AstraPEP.
+
+ACT treatments commonly involve around 10 sets of 10 exhalation breaths into an ACT device.
+Participants generally used 1 of 4 different ACT devices compatible with the pressure sensor; the Acapella, Aerobika, PariPEP or AstraPEP.
+Each device produced breaths with a slightly different overall shape, e.g.
+with small oscillations.
+Sometimes participants used different devices over different days.
+Each set helps to loosen mucus, and between sets, patients normally 'huff' and cough to expel it.
+The treatments are usually performed one or more times per day depending on how well the person is or how much mucus they normally produce; participants did not have identical prescriptions.
+For the study, the ACT device was fitted with a bespoke wireless pressure sensor sampling pressure at 10~Hz, recording pressure values in time-stamped millibars.
+
 \subsubsection{Processing breath data}
+
 Several pre-processing steps were performed on the data which were a function of the custom sensor used: including assembling data packets and removing duplicate values.
@@
 Because of the asymmetrical nature of the exhalation peaks and sparsity of the signal, we used a sparsity-based de-noising approach with an asymmetry factor, called BEADS \cite{Ning2014}, which is designed for use with chromatogram waveforms which have a similar form.
-% figure 1
+An example of pre-processing raw ACT data is shown in figure \ref{fig:ACTpreprocessing}.
+
+\begin{figure}[htb]
+  \includegraphics[width=\textwidth]{ACTpreprocessing.png}
+  \centering
+   \caption{Left: raw ACT data in two packets, with baseline drift.
+Right: Data packets assembled into a treatment and baseline drift removed.
+This treatment shows 11 sets of 10 exhalations (the positive peaks), lasting 33 minutes.}
+  \label{fig:ACTpreprocessing}
+\end{figure}
+
 In total, we recorded 5x,000 sensor readings.
-This includes data from every time the ACT sensor was turned on, including for testing and demonstration.
+This included data from every time the ACT sensor was turned on, including for testing and demonstration.
 Once we removed readings shorter than 15 seconds, treatments with no breath exhalations, and situations with sensor misuse/malfunction, we were left with 2x,000 ACT treatments for analysis.
@@
 \subsubsection{Featurising breath data}
-% figure 2
-The defining characteristics of ACT treatments were summarised in a range of features (variables): breath count, breath duration, breath amplitude, and treatment duration.
-In order to derive these features, our breath identification algorithm recognises breaths composed of multiple peaks, and having breath shapes which are characteristic of the 3 different ACT devices used in the study.
+
+A typical ACT waveform which has been pre-processed, is shown in figure \ref{fig:individual_breaths}.
+The defining characteristics of ACT treatments were summarised in a range of features (variables) including: breath count, breath duration, breath amplitude, and treatment duration.
+In order to derive these features, our breath identification algorithm recognises not only simple breaths, but complex breaths composed of multiple peaks, and having a variety of shapes & oscillations which are characteristic of the 3 different ACT devices used in the study.
 In addition to mean breath amplitude, we use amplitude standard deviation as a feature, in order to capture inconsistency or disorderliness in physiotherapy.
@@
 
+\begin{figure}[htb]
+  \includegraphics[width=\textwidth]{individual_breaths.png}
+  \centering
+   \caption{A typical, pre-processed ACT waveform. This treatment is 20 seconds long and contains 7 exhalation breaths.}
+  \label{fig:individual_breaths}
+\end{figure}
+
 The aforementioned features capture the characteristics of existing treatments.
 However it is common for prescribed treatments to be skipped, and days with no treatments is an important signal itself.
-Several features were derived which capture missing data: treatment adherence to date, skipped days since last treatment, rolling adherence over 7 days.
+Several features were derived which captured missing data: treatment adherence to date, skipped days since last treatment, rolling adherence over 7 days.
 These features are on a per-day level, as participants are prescribed 1 or 2 treatments per day.
@@
 \subsubsection{Featurising physical activity data}
-The Fitbit activity tracker (Fitbit AltaHR) provides each participant’s step count per minute.
-While real-time heartbeat data would allow for a detailed analysis of heartrate waveforms, the Fitbit only provides average heartrate, with a variable sampling frequency ranging from 3 to 12 values per minute.
-To achieve consistent frequency and comparable values, the heartrate values are averaged for each minute.
+The Fitbit activity tracker (Fitbit AltaHR) provided each participant’s step count per minute.
+While real-time heartbeat data would allow for a detailed analysis of heart rate waveforms, the Fitbit only provides average heart rate, with a variable sampling frequency ranging from 3 to 12 values per minute.
+To achieve consistent frequency and comparable values, the heart rate values were averaged for each minute.
+An example of heart rate data is shown in figure \ref{fig:HR}.
 
-Mindful of these data limitations, we extract high-level patterns from the minute averaged waveforms at a day level.
-The heartrate and steps datasets are featurised using summary statistics per day using counts, frequency and interval values such as number of minutes in moderate to vigorous intensity physical activities (MVPA) (heartrate above 120 bpm) per day, step cadence (number of steps/minutes of wear) at day level, and number of hours with more than 500 steps at day level.
+\begin{figure}[htb]
+  \includegraphics[width=\textwidth]{HR.png}
+  \centering
+   \caption{One day showing heartrate data with a sustained period of heartrate in MVPA.}
+  \label{fig:HR}
+\end{figure}
 
-% figure 3
+Mindful of these data limitations, we extracted high-level patterns from the minute-averaged waveforms at a day level.
+The heart rate and steps data sets were featurised using summary statistics per day using counts, frequency and interval values, such as number of minutes in moderate to vigorous intensity physical activities (MVPA) (heart rate above 120 bpm) per day, step cadence (number of steps/minutes of wear) per day, and number of hours with greater than 500 steps per day.
 
-\subsubsection{Missing activity data}
-
-To account for missing data during times that the Fitbit is not worn, the wear percent of the Fitbit is calculated from the heartrate values.
-This measure is used to select only days with a substantial portion of data for the day (80\% of wear time from 6 am to 6 pm) for analysis.
-The wear time is also used to normalise some features and to obtain step cadence for minutes where heartrate data is present.
+To account for missing data during times that the Fitbit was not worn, the wear percent of the Fitbit was calculated from the valid heart rate values.
+This measure was used to select only days with a substantial portion of data for the day (80~\% of wear time from 6~am to 6~pm), resulting in a dataset containing 9760 days used for analysis.
+The wear time was also used to normalise some features and to obtain step cadence for minutes where heart rate data was present.
 
@@
 We use the Hopkins statistic [ref] to measure clustering tendency.
-Features with a Gaussian distribution are normalised to have mean =0 and standard deviation 1.
-In order to visualise the results, we use dimensionality reduction.
-We use the first two principle components from principal component analysis [ref], Uniform Manifold Approximation and Projection (UMAP), and t-Distributed Stochastic Neighbor Embedding (T-SNE).
-We identify outliers using xxx methods and remove less than 2\%, with these cases being recognised as device malfunctions or unusual human behaviour.
+Features with a Gaussian distribution were normalised to have mean~=~0 and standard deviation~=~1.
+In order to visualise the results, we used dimensionality reduction.
+We used the first two principle components from principal component analysis [ref], Uniform Manifold Approximation and Projection (UMAP) [ref], and t-Distributed Stochastic Neighbor Embedding (T-SNE [ref].
+We identified outliers using xxx (Olga) methods and removed less than xxx2~\%, with these cases being recognised as device malfunctions or abnormal behaviour.
 
-We employ the following clustering methods: k-means, Density-based spatial clustering of applications with noise (DBSCAN) and hierarchical clustering.
+We employed the following clustering methods: k-means, Density-based spatial clustering of applications with noise (DBSCAN) [ref?] and hierarchical clustering.
 
 \section{Results}
+
 \subsection{Clustering breath waveforms}
+
 \subsubsection{Feature selection for clustering}
-Some features were excluded from clustering, but provide interesting physiotherapy insights.  
-The number of sets in a treatment were expected to be very influential in clustering, as a regular pattern of sets shows close adherence to physiotherapy advice.
-However we found that less than 20xx\% of treatments had discernible sets, so this feature could not be used
 
-% table 1
+Some features were excluded from clustering, but provided interesting physiotherapy insights.  
+The number of sets in a treatment were expected to be influential in clustering, as a regular pattern of sets would show close adherence to physiotherapy advice.
+However we found that less than 20xx\% (Tempest) of treatments had discernible sets, so this feature could not be used.
+The features used for clustering are shown in table X below, while the full list of features is shown in appendix X.
 
-\subsubsection{ACT clustering results}
-The results of the four techniques used for clustering are show in table x. K-means was found to perform best because we liked the way..… K-means is a simple
+\subsubsection{Comparison of clustering techniques}
+We compared the three clustering techniques (see appendix) and found the best performance with k-means.
 
-% table 2
-% table 3
-% figure 4
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[width=\textwidth]{featureTables.png}
+  \caption{To be replaced with a proper table.}
+\end{figure}
+
+\subsubsection{ACT K-means clustering results}
+The purpose of clustering into sub-types is to summarise how a participant is performing their ACT.
+We chose k=4 clusters to maximise both silhouette value (average 0.18), and the interpretability of the results.
+If k is too low, the sub-types are not granular enough, if k is too high it becomes challenging to remember the characteristics of each sub-type.
+
+The centroid values of the 4 clusters are shown in table X below. Based on these centroids, the four ACT sub-types, which we named alpha, beta, gamma and delta have the following interpretations:
+
+\begin{itemize}
+\item Alpha - Low adherence  ....
+\item Beta - High inconsistency ....
+\item Gamma -  ....
+\item Delta -  ....
+\end{itemize}
 
@@
 
-We have checked the clustering stability on different data partitions.
-Random sampling of the data (80\% of data) and using k-means clustering provided very similar results comparing to clustering on the full dataset:  shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index  (2200 v 1770 respectively) are comparable.
-% figure
-We also have done validation of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters.
-Figure xx shows the same patient trajectory for the Clustering A and Clustering B.
-Clustering B uses breath amplitude features (mean and standard dev) as well as breathRollWeekAdherenceScore in addition to features that Clustering A uses.
-Notable that  once computer gaming is introduced to ACT in both cases A and B the same patient moves from cluster with “expected” breath counts per ACT to cluster characterized by much bigger breath counts.
-This validates that there is a change in patient’s behaviour that we’ve captured.
-Additional sdBreathAmplitude feature make results more explainable from clinical point of view as original cluster for the patient in Clustering B was characterized by centroid with the biggest sdBreathAmplitude value.
-% figure
+The ACT sub-types should be sensitive enough to capture changes in ACT, particularly during interventions.
+Figure ~\ref{fig:act_res} shows that clusters are sensitive enough to track changes in a participants' adherence through different study interventions, which validates our approach.
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[]{fig_ACT_clust}
+  \caption{the four ACT clusters overlaid with a single participant (black), showing that the participant’s cluster membership changes over time.
+Clustering results are visualised in 2-dimensions using PCA (left) and UMAP (right).
+The time points are before ACT feedback (number of registered breaths) is introduced (top), after ACT feedback is introduced (middle), and once computer gaming is introduced to ACT routine (bottom).
+Each point represents one treatment of one participant.
+}
+  \label{fig:act_res}
+\end{figure}
+
+We also assessed the clustering stability on different data partitions.
+Random sampling 80~\% of the data and clustering using k-means provided very similar results with clustering on the full dataset: figure \ref{fig:ACTclusteringon80} shows that the shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index (2200 v 1770 respectively) were comparable.
+
+\begin{figure}[!htb]
+  \centering
+  \caption{K-means clustering results on full dataset (left) and subset of data (right).}
+  \includegraphics[]{ACTclusteringOn80.png}
+  \label{fig:ACTclusteringon80}
+\end{figure}
+
+*Not sure about this part. Maybe move to appendix?
+We also assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters.
+Figure \ref{fig:ACTclusteringAB} shows the same patient trajectory for clustering with feature selection A and clustering with feature selection B, which includes two additional features: breath amplitude standard deviation, and rolling weekly adherence score for breath count.
+It is notable that once computer gaming was introduced to ACT in both cases A and B the same patient moved from cluster with “expected” breath counts per ACT to a cluster characterized by much higher breath counts.
+This validates that there was a change over time in the patient’s ACT pattern that could be captured.
+The additional sdBreathAmplitude feature make results more explainable from a clinical point of view, as the original cluster for the patient in Clustering B was characterized by a centroid with the highest sdBreathAmplitude value.
+
+\begin{figure}[!htb]
+  \centering
+  \caption{Patient Trajectory on set A of features (without breath amplitude) and set B of features.}
+  \includegraphics[]{ACTclusteringAB.png}
+  \label{fig:ACTclusteringAB}
+\end{figure}
 
@@
 
-Clustering of physical activity data was used to extract daily patterns from the steps and heartrate waveforms that can be utilised to individually characterise the physical activity levels and patterns of individual patients.
+The purpose of clustering into sub-types is to summarise each participant's physical activity data.
+Clustering of physical activity data was used to extract daily patterns from the steps and heart rate waveforms and characterise the physical activity levels and patterns of individual patients using a dataset containing 9760 individual days.
 The different clusters should extract more complex characteristics than the ones which could be obtained by simply combining some of the extracted features.
@@
 
-The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, the mean of total step count per hour, the step cadence for the periods in MVPA and the number of hours in the day with 0 < steps count < 500.
+The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, the mean of total step count per hour, the step cadence for the periods in MVPA and the number of hours in the day with 0~< ~steps~count~<~500.
 Features were selected as a result of feature selection, correlation and outlier analysis.
-Four clusters were found by Kmeans with the following characteristics:
+Three sub-types were desired for the steps clusters covering small, medium and large amounts of steps activity throughout the day.
+Four clusters (as shown in figure \ref{fig:stepsClusters}) were chosen in this case to cover these three sub-types as well as a sub-type for the instances with missing data (XS cluster).
 
-\begin{itemize}
-\item XS - this cluster contains waveforms with very little steps activity or much missing step data 
-\item S - moderate step activity and consistent cadence throughout the day with very few periods of high step activity 
-\item M - consistent moderate to high step rate across the day with period of sustained high step activity  
-\item L - High step rate throughout the day with longer periods of sustained high step activity
-\end{itemize}
-%figure
-%table
+\begin{figure}[htb]
+  \centering
+  \caption{Steps clusters displayed on the first two principle components.}
+  \includegraphics[]{stepsClusters.png}
+  \label{fig:stepsClusters}
+\end{figure}
 
-subsubsection{Heartrate clusters}
+\begin{table}[H]
+  \caption{Steps Clusters Interpretation}
+  \label{steps_inter}
+  \centering
+  \begin{tabular}{ c|l }
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster name & Cluster Interpretation \\
+    \hline
+    XS & Waveforms with very little steps activity or high amount of missing data \\
+    \hline
+    S & Moderate step activity, consistent cadence with few periods of high activity\\
+    \hline
+    M & Consistent moderate to high step rate with a period of sustained high step activity \\
+    \hline
+    L & High step rate with longer periods of sustained high step activity \\
+    \bottomrule
+    \end{tabular}
+\end{table}
 
-The number of minutes with heartrate higher than 120, coefficient of variation for heartrate values at a day level, number of time heartrate goes above 100 and 120 and the median average hourly heartrate were used as features for the heartrate clusters.
-Five clusters were found by Kmeans with the following chacteristics:
-\begin{itemize}
-\item S - Small heartrate values and consistently below 120 
-\item M - Consistent variations around 100 with very few spikes in heartrate values 
-\item M1 - Waveform with high variability and periods with spikes above 120 and fluctuations above the 120 mark 
-\item M2 - Consistent variations around 100 with spikes in values above 120 
-\item L - Large values around and above 120 
-\end{itemize}
-%figure
-%table
+\subsubsection{Heart rate clusters}
+
+The following were used as features for heart rate clustering: the number of minutes with heart rate higher than 120, coefficient of variation for heart rate values per day, number of time heart rate exceeds 100 and 120, and the median average hourly heart rate.
+Experiments were carried out with 3, 4 and 5 clusters and the choice of k was based on the cluster interpretability results.
+The sub-types found when k=5 were the best at identifying more complex patterns and grouping similar waveforms together.
+The five clusters found by K-means (as shown in figure \ref{fig:HRclusters} have the following interpretation:
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[]{HRclusters.png}
+  \caption{PCA of representation of the five heart rate clusters. }
+  \label{fig:HRclusters}
+\end{figure}
+
+\begin{table}[H]
+  \caption{Heartrate Clusters Interpretation}
+  \label{hr_inter}
+  \centering
+  \begin{tabular}{ c|l }
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster name & Cluster Interpretation \\
+    \hline
+    S & Small heart rate values and heart rate consistently below 120 \\
+    \hline
+    SM & Consistent variations around 100 with very few spikes in heart rate values \\
+    \hline
+    M1 &  Waveform with high variability and periods with spikes above 120 \\
+    \hline
+    M2 & Consistent variations around 100 with spikes in values above 120 \\
+    \hline
+    L &  Large values around and above 120 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
 
@@
 
-The steps and heartrate clusters were validated with the clinical experts.
+The steps and heart rate clusters were validated with clinical experts.
+The clusters were found to be meaningful and useful ways to categorise the waveforms.
 Individual waveforms were manually inspected together with the cluster boundaries to ensure consistency across cluster interpretations and the actual cluster characteristics.
 Additionally, other features that were not included in the clustering analysis were used to validate the clusters and their interpretations.
-Certain expected conditions were checked against the cluster labels and plotted using the PCA graph such as Figure Xxx below.
+Certain expected conditions were checked against the cluster labels and plotted using the PCA graph such as figure \ref{fig:FitbitClusterValida} below.
+Cluster boundaries for both steps and heartrate clusters were also utilised to validate the cluster properties and interpretation, these can be found in the Appendix.
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[]{FitbitClusterValidation.png}
+  \caption{: Crosses are days with daily cadence above 30 which are all part of the L cluster.}
+  \label{fig:FitbitClusterValidation}
+\end{figure}
 
@@
 
-Project Fizzyo involves collecting novel data about how CF physiotherapy advice, ACTs and physical activity, is adhered to.
-In close collaboration with CF experts, we develop methods to interrogate this data.
-We derive clinically meaningful features from ACT and activity (heartrate and steps) waveforms.
-We use unsupervised clustering to find sub-types of this data.
-
-Project Fizzyo is an ongoing study, with participation lasting 18 months.
-7 months into the trial we do not draw any conclusions, as we do not have an equal number of days of data from patients.
-Instead, we present a framework for analysis, which will empower clinicians to interpret the impact of interventions such as gamification, and what type of physiotherapy leads to better clinical outcomes.
+Project [Anonymised] involves collecting novel data about how CF physiotherapy advice on ACTs and physical activity is adhered to in everyday life, information which has never been available to physiotherapists.
+In close collaboration with CF experts, we developed methods to interrogate and make sense of this data.
+Project [Anonymised] is an ongoing study, with participation lasting 18 months.
+8 months into the trial we do not draw any conclusions, as we do not have an equal number of days of data from patients recruited over 8 months.
+Instead, we present a framework for analysis, which will empower clinicians to understand how physiotherapy advice is actually adhered to, interpret the impact of interventions such as gamification when the trial is complete., and eventually understand what type of physiotherapy leads to better clinical outcomes.
+Our framework involves processing the ACT and Fitbit data, deriving clinically meaningful features from it which characterise this novel data, and then using unsupervised clustering to find sub-types of this data.
+The clusters provide a useful way of tracking participants' trajectories across sub-types over time.
 
@@
 \subsection{Figures}
-
-\begin{figure}
-  \centering
-  \caption{Sample figure caption.}
-  \includegraphics[]{sample_fig}
-\end{figure}
 
@@
     Col1 & Col2 & Col3 \\
-    \midrule
+    \hline
     a & 0 & z\\
@@
 \appendix
+
 \section{Appendix : Table of physical activity data features}
 
-\begin{table}[h]
-  \caption{Physical activity features}
-  \label{fitbit_features}
-  \centering
-  \begin{tabular}{ |l|l|l| }
+\begin{longtable}{  n|d }
 \toprule
 \cmidrule(r){1-2}
-    Feature Name & Feature description & Waveform extracted from \\
-    \midrule
-    userif & 1099 & 0.95 \\
+\textbf{Feature Name} & \textbf{Feature description} \\
+\hline
+\endfirsthead
+Minutes wear & Number of minutes in a day with heartrate value present \\
+\hline
+Percentage of wear & $\frac{\text{Minutes wear}}{24\cdot60}\cdot100\%$\\
+\hline
+Wear during sleep & Indicator value: 1,  worn during sleep or 0, otherwise \\
+\hline
+Active hours & Number of hours with a step count > 500  \\
+\hline
+Low active hours & Number of hours with a step count < 500  \\
+\hline
+Mean hourly steps & Average of hourly step count \\
+\hline
+Step count & Total step count per day \\
+\hline
+Normalised step count & Rate of steps per minute when fitbit is worn \\
+ & $\frac{Step Count}{Minutes Worn}$ \\
+\hline
+ & Number of minutes with step count > 100 with the following constraint: \\
+Active Minutes Step 0 & - No minimum of consecutive minutes\\
+Active Minutes Step 2 & - At least 2 consecutive minutes > 100 \\
+Active Minutes Step 5 & - At least 5 consecutive minutes > 100 \\
+Active Minutes Step 10 & - At least 10 consecutive minutes > 100 \\
+Active Minutes Step 20 & - At least 20 consecutive minutes > 100 \\
+\hline
+Active minutes Step for the active period & Number of active minutes with step > 100 from 6am to midnight \\
+\hline
+Coefficient of Variation Step & Coefficient of variation for steps (removing zero values) at day level \\
+ & $CoV = \frac{Standard Deviation}{Mean} * 100$ \\
+ \hline
+ Normalised step count for periods of : & \\
+ - moderate activity & Rate of steps per minute when the heartrate is between 120 and 140. \\ 
+ - vigorous activity & Rate of steps per minute when the heartrate is above 140. \\
+ - moderate and vigorous activity & Rate of steps per minute when the heartrate is above 120. \\
+ - non-moderate and non-vigorous activity & Rate of steps per minute when the heartrate is below 120.\\
+ \hline
+ Proxy for Resting Heartrate & Mean of the lowest 5 heartrate minutes averages \\
+ \hline
+ & Number of minutes with heartrate > 120 with the following constraint: \\
+Active Minutes Heartrate 0 & - No minimum of consecutive minutes\\
+Active Minutes Heartrate 2 & - At least 2 consecutive minutes > 120 \\
+Active Minutes Heartrate 5 & - At least 5 consecutive minutes > 120 \\
+Active Minutes Heartrate 10 & - At least 10 consecutive minutes > 120 \\
+Active Minutes Heartrate 20 & - At least 20 consecutive minutes > 120 \\
+\hline
+Active minutes Heartrate for the active period & Number of active minutes with heartrate > 100 from 6am to midnight \\
+\hline
+Coefficient of Variation Heartrate & Coefficient of variation for heartrate (removing zero values) at day level \\
+ & $CoV = \frac{Standard Deviation}{Mean} * 100$ \\
+ \hline
+ Number of threshold switches above 100 & Number of times the heartrate goes above 100 from below this threshold at day level. \\ 
+ Number of threshold switches above 120 & Number of times the heartrate goes above 120 from below this threshold at day level. \\ 
+ Number of threshold switches above 140 & Number of times the heartrate goes above 140 from below this threshold at day level. \\ 
+ \hline
+ & Number of minutes with step count > 100 and heartrate > 120 with the following constraint: \\
+Active Minutes Step and Heartrate 0 & - No minimum of consecutive minutes\\
+Active Minutes Step and Heartrate 2 & - At least 2 consecutive minutes \\
+Active Minutes Step and Heartrate 5 & - At least 5 consecutive minutes \\
+Active Minutes Step and Heartrate 10 & - At least 10 consecutive minutes \\
+Active Minutes Step and Heartrate 20 & - At least 20 consecutive minutes \\
+\hline
+Heartrate Hourly max, min and median & Maximum. minimum and median of average hourly heartrate. \\
 \bottomrule
-    \end{tabular}
-\end{table}
+\end{longtable}
 
-\section{Appendix : Cluster metrics for physical activity data}
+\section{Appendix : Table of ACT treatment features}
+
+\begin{longtable}{ n|d}
+\toprule
+\cmidrule(r){1-2}
+\textbf{Feature Name} & \textbf{Feature description} \\
+\hline
+Breath Count & Number of breaths the person has made though an ACT  \\ 
+ ~ & device during a treatment. A breath is counted if its maximum amplitude is \\ ~ & above the threshold of 8 mm H2O.\\
+ \hline
+ Mean Breath Amplitude & Average breath amplitude (in mm H2O) during an ACT treatment. \\
+ Minimum Breath Amplitude & Similarly minimum, maximum and standard deviation of breath amplitude. \\
+ Maximum Breath Amplitude & ~ \\
+ Standard Deviation of Breath Amplitude & ~ \\
+ \hline
+ Mean Breath Duration & Average duration (in seconds) of breath during an ACT treatment. \\ 
+ Minimum Breath Duration &  Similarly minimum, maximum and standard deviation of \\ 
+ Maximum Break Duration & breath duration in a treatment. \\
+ Standard Deviation of Breath Duration & ~ \\
+ \hline
+ Break Count & Number of breaks (between breaths) in a treatment. \\
+ \hline
+ Mean Break Duration & Average duration (in seconds) of breaks between breaths. \\ 
+ Minimum Break Duration & Similarly minimum, maximum and standard deviation of \\ 
+ Maximum Break Duration & breaks duration in a treatment. \\
+ Standard Deviation of Break Duration & ~ \\
+ \hline
+ Set Count & Number of sets in a treatment. Prescribed treatment may have up to 10 \\
+ ~ & sets. Set is a group of breaths separated by huffing and coughing activity. \\
+ \hline
+ Pressure Readings Count & Number of pressure reading registered by ACT device during a treatment. \\
+ ~ & Used to filter out "test" device sessions that are not part of a treatment. \\
+ \hline
+ Pressures Readings Completeness Score & Data completeness score for the day \\
+ ~ & If a treatment has less than 150 pressure readings it's completeness \\
+ ~ & score is 0. \\
+ \hline
+ Treatment Duration & Treatment duration (in minutes). \\
+ \hline
+ Treatment Daily Adherence Score & Number of completed treatments out of the prescribed ones (for the \\
+ ~ & given day). \\
+ \hline
+ Breath Count Daily Adherence Score & Number of registered breaths across all treatments for the given day \\
+ ~ & out of the prescribed amount of breaths. \\ 
+ \hline
+ Treatment Rolling Weekly Adherence Score & Number of completed treatments during the last 7 days (for a given date) \\
+ ~ & out of the prescribed number of treatments for this period. \\
+ \hline
+ Breath Count Rolling Weekly Adherence Score & Number of registered breaths across all treatments during the last 7 days \\
+ ~ & (for a given date) out of the prescribed amount of breaths for this period. \\ 
+ \hline
+ Treatment Weekly Adherence Score & Number of completed treatments during the last calendar week (for a given date) \\
+ ~ & out of the prescribed number of treatments for this period. \\
+ \hline
+ Breath Count Weekly Adherence Score & Number of registered breaths across all treatments during the last calendar week \\
+ ~ & (for a given date) out of the prescribed amount of breaths for this period.\\
+ \hline
+ Cumulative Adherence Score & Cumulative adherence score to date. \\
+ ~ & How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount. \\
+ \hline
+ Breathing Time Percentage & Time in a treatment spent on an exhalation, vs not.\\
+ \hline
+ Skipped Days Count & Skipped days since the last treatment. \\
+ \bottomrule
+\end{longtable}
+
+\section{Appendix : Cluster results and metrics for physical activity data}
 
@@
 
-\textbf{Steps clusters}
+\subsection{Steps clusters}
 
+\subsubsection{K-Means Clustering}
+
+\paragraph{Cluster boundaries}
+Cluster boundaries for the four steps clusters can be seen in Figure \ref{fig:stepsBoundaries}.
+
+\begin{figure}[H]
+  \centering
+  \caption{Steps Cluster Boundaries}
+  \includegraphics[scale=0.5]{steps_cluster_boundaries.png}
+  \label{fig:stepsBoundaries}
+\end{figure}
+
+\paragraph{Cluster Metrics}
 Calinski-harabasz index for the steps clusters is 6021.14 .
 
-\begin{table}[h]
-  \caption{Steps clusters Metrics}
+\begin{table}[H]
+  \caption{Steps clusters Metrics Kmeans}
   \label{steps_metrics}
   \centering
-  \begin{tabular}{ |l|c|c|}
+  \begin{tabular}{ c|c|c}
     \toprule
@@
     Cluster name & Cluster Size & Average Silhouette width \\
-    \midrule
+    \hline
     XS & 1099 & 0.95 \\
@@
 
-\textbf{Heartrate clusters}
+\subsubsection{Hierarchical Clustering with complete linkage}
 
+The clustering results using Hierarchical clustering with complete linkage on steps data can be seen in Figure \ref{fig:stepsClustersHC}.
+
+\begin{figure}[htb]
+  \centering
+  \caption{Steps clusters displayed on the first two principle components obtained using Hierarchical clustering}
+  \includegraphics[]{steps_HC_results.png}
+  \label{fig:stepsClustersHC}
+\end{figure}
+
+Calinski-harabasz index for the steps clusters is 1518.1865 .
+
+\begin{table}[H]
+  \caption{Steps clusters Metrics HC}
+  \label{steps_metrics}
+  \centering
+  \begin{tabular}{ c|c|c}
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster number & Cluster Size & Average Silhouette width \\
+    \hline
+    1 & 4996 & 0.12 \\
+    2 & 4056 & 0.11 \\
+    3 & 677 & 0.18 \\
+    4 & 31 & 0.41 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\subsection{Heartrate clusters}
+
+\subsubsection{K-Means Clustering}
+\paragraph{Cluster boundaries}
+Cluster boundaries for the five heartrate clusters can be seen in Figure \ref{fig:hrBoundaries}.
+
+\begin{figure}[H]
+  \centering
+  \caption{Heartrate Cluster Boundaries}
+  \includegraphics[scale=0.5]{hr_cluster_boundaries.png}
+  \label{fig:hrBoundaries}
+\end{figure}
+
+\paragraph{Cluster Metrics}
 Calinski-harabasz index for the heartrate clusters is 5090.37 .
 
-\begin{table}[h]
-  \caption{Heartrate clusters Metrics}
+\begin{table}[H]
+  \caption{Heartrate clusters Metrics Kmeans}
   \label{hr_metrics}
   \centering
-  \begin{tabular}{ |l|c|c|}
+  \begin{tabular}{ |c|c|c|}
     \toprule
@@
     Cluster name & Cluster Size & Average Silhouette width \\
-    \midrule
+    \hline
     S & 1246 & 0.28 \\
@@
 
+\subsubsection{Hierarchical Clustering with complete linkage}
+
+The clustering results using Hierarchical clustering with complete linkage for the heartrate data can be seen in Figure \ref{fig:hrClustersHC}.
+
+\begin{figure}[htb]
+  \centering
+  \caption{Heartrate clusters displayed on the first two principle components obtained using Hierarchical clustering}
+  \includegraphics[]{hr_HC_results.png}
+  \label{fig:hrClustersHC}
+\end{figure}
+
+Calinski-harabasz index for the heartrate clusters is 135.51 .
+
+\begin{table}[H]
+  \caption{Heartrate clusters Metrics HC}
+  \label{hr_metrics}
+  \centering
+  \begin{tabular}{ |c|c|c|}
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster number & Cluster Size & Average Silhouette width \\
+    \hline
+    1 & 4221 & -0.07 \\
+    2 & 4901 & -0.06 \\
+    3 & 607 & -0.10 \\
+    4 & 24 & 0.07 \\
+    5 & 7 & -0.06 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
 \end{document}

### 2019-09-05 draft.tex
ACT: Kmeans metrics and interpretation

@@
     Cluster name & Cluster Interpretation \\
-    \hline
+    \midrule
     XS & Waveforms with very little steps activity or high amount of missing data \\
-    \hline
+    \midrule
     S & Moderate step activity, consistent cadence with few periods of high activity\\
-    \hline
+    \midrule
     M & Consistent moderate to high step rate with a period of sustained high step activity \\
-    \hline
+    \midrule
     L & High step rate with longer periods of sustained high step activity \\
@@
     Cluster name & Cluster Interpretation \\
-    \hline
+    \midrule
     S & Small heart rate values and heart rate consistently below 120 \\
-    \hline
+    \midrule
     SM & Consistent variations around 100 with very few spikes in heart rate values \\
-    \hline
+    \midrule
     M1 &  Waveform with high variability and periods with spikes above 120 \\
-    \hline
+    \midrule
     M2 & Consistent variations around 100 with spikes in values above 120 \\
-    \hline
+    \midrule
     L &  Large values around and above 120 \\
@@
 
-\section{Usage of tables and figures}
-\subsection{Figures}
-
-\subsection{Tables}
-\begin{table}
-  \caption{Sample table title}
-  \label{sample_table}
-  \centering
-  \begin{tabular}{lll}
-    \toprule
-    \cmidrule(r){1-2}
-    Col1 & Col2 & Col3 \\
-    \hline
-    a & 0 & z\\
-    b & 1 & y \\
-    c & 2 & x \\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Here is how you can reference a table \ref{sample_table}
-
 \bibliographystyle{abbrv}
@@
 
-\begin{longtable}{  n|d }
+\begin{longtable}{  n | d }
 \toprule
 \cmidrule(r){1-2}
-\textbf{Feature Name} & \textbf{Feature description} \\
-\hline
+\textbf{Fitbit features} & \textbf{Feature description} \\
+\midrule
 \endfirsthead
 Minutes wear & Number of minutes in a day with heartrate value present \\
-\hline
-Percentage of wear & $\frac{\text{Minutes wear}}{24\cdot60}\cdot100\%$\\
-\hline
+\midrule
+Percentage of wear & $\frac{\textit{Minutes wear}}{24\cdot60}\cdot100\%$\\
+\midrule
 Wear during sleep & Indicator value: 1,  worn during sleep or 0, otherwise \\
-\hline
+\midrule
 Active hours & Number of hours with a step count > 500  \\
-\hline
+\midrule
 Low active hours & Number of hours with a step count < 500  \\
-\hline
+\midrule
 Mean hourly steps & Average of hourly step count \\
-\hline
+\midrule
 Step count & Total step count per day \\
-\hline
-Normalised step count & Rate of steps per minute when fitbit is worn \\
- & $\frac{Step Count}{Minutes Worn}$ \\
-\hline
- & Number of minutes with step count > 100 with the following constraint: \\
-Active Minutes Step 0 & - No minimum of consecutive minutes\\
-Active Minutes Step 2 & - At least 2 consecutive minutes > 100 \\
-Active Minutes Step 5 & - At least 5 consecutive minutes > 100 \\
-Active Minutes Step 10 & - At least 10 consecutive minutes > 100 \\
-Active Minutes Step 20 & - At least 20 consecutive minutes > 100 \\
-\hline
-Active minutes Step for the active period & Number of active minutes with step > 100 from 6am to midnight \\
-\hline
-Coefficient of Variation Step & Coefficient of variation for steps (removing zero values) at day level \\
- & $CoV = \frac{Standard Deviation}{Mean} * 100$ \\
- \hline
- Normalised step count for periods of : & \\
- - moderate activity & Rate of steps per minute when the heartrate is between 120 and 140. \\ 
- - vigorous activity & Rate of steps per minute when the heartrate is above 140. \\
- - moderate and vigorous activity & Rate of steps per minute when the heartrate is above 120. \\
- - non-moderate and non-vigorous activity & Rate of steps per minute when the heartrate is below 120.\\
- \hline
- Proxy for Resting Heartrate & Mean of the lowest 5 heartrate minutes averages \\
- \hline
- & Number of minutes with heartrate > 120 with the following constraint: \\
-Active Minutes Heartrate 0 & - No minimum of consecutive minutes\\
-Active Minutes Heartrate 2 & - At least 2 consecutive minutes > 120 \\
-Active Minutes Heartrate 5 & - At least 5 consecutive minutes > 120 \\
-Active Minutes Heartrate 10 & - At least 10 consecutive minutes > 120 \\
-Active Minutes Heartrate 20 & - At least 20 consecutive minutes > 120 \\
-\hline
-Active minutes Heartrate for the active period & Number of active minutes with heartrate > 100 from 6am to midnight \\
-\hline
-Coefficient of Variation Heartrate & Coefficient of variation for heartrate (removing zero values) at day level \\
- & $CoV = \frac{Standard Deviation}{Mean} * 100$ \\
- \hline
- Number of threshold switches above 100 & Number of times the heartrate goes above 100 from below this threshold at day level. \\ 
- Number of threshold switches above 120 & Number of times the heartrate goes above 120 from below this threshold at day level. \\ 
- Number of threshold switches above 140 & Number of times the heartrate goes above 140 from below this threshold at day level. \\ 
- \hline
- & Number of minutes with step count > 100 and heartrate > 120 with the following constraint: \\
-Active Minutes Step and Heartrate 0 & - No minimum of consecutive minutes\\
-Active Minutes Step and Heartrate 2 & - At least 2 consecutive minutes \\
-Active Minutes Step and Heartrate 5 & - At least 5 consecutive minutes \\
-Active Minutes Step and Heartrate 10 & - At least 10 consecutive minutes \\
-Active Minutes Step and Heartrate 20 & - At least 20 consecutive minutes \\
-\hline
+\midrule
+Normalised step count & Rate of steps per minute when fitbit is worn $\frac{\textit{Step Count}}{\textit{Minutes Worn}}$ \\
+\midrule
+ Step active minutes $k$, for $k\in \{0,2,5,10,20\}$ & Number of active minutes in a day.
+A minute is considered active if it has a step count > 100 and is part of a sequence of at least $k$ consecutive minutes with step count > 100.
+Note that for $k=0$ there is no restriction on the number of consecutive minutes\\
+\midrule
+Step active minutes during waking period & Number of minutes with step > 100 from 6am to midnight \\
+\midrule
+Step coefficient of variation & Coefficient of variation of minute step counts $CoV = \frac{\sigma (\textit{Step count})}{\mu (\textit{Step count})} \cdot 100$ \\
+ \midrule
+ Normalised step count for periods of : &  Rate of steps per minute when the heartrate is: \\
+ - moderate activity & - between 120 and 140. \\ 
+ - vigorous activity & - above 140. \\
+ - moderate and vigorous activity & - above 120. \\
+ - non-moderate and non-vigorous activity & - below 120.\\
+ \midrule
+ Resting heartrate proxy & Mean of the 5 minutes with the lowest heartrate average \\
+ \midrule
+Heartrate active minutes $k$, for $k\in \{0,2,5,10,20\}$ & Number of active minutes in a day.
+A minute is considered active if it has heartrate > 120 and is part of a sequence of at least $k$ consecutive minutes with heartrate > 120.\\
+\midrule
+Heartrate active minute during waking period & Number of minutes with heartrate > 120 from 6am to midnight \\
+\midrule
+Heartrate coefficient of variation & Coefficient of variation for heartrate minute averages $CoV = \frac{\sigma (\textit{Heartrate})}{\mu (\textit{Heartrate})} \cdot 100$ \\
+ \midrule
+ Number of switches above threshold $t$, for $t\in \{100,120,140\}$ & Number of times the heartrate goes above the threshold $t$ from below this threshold.\\ 
+ \midrule
+ Step and heartrate active minutes $k$, for $k\in \{0,2,5,10,20\}$ & Number of active minutes in a day.
+A minute is considered active if it has a step count > 100 and heartrate > 120 and is part of a sequence of at least $k$ consecutive minutes which satisfy both conditions.
+\midrule
 Heartrate Hourly max, min and median & Maximum. minimum and median of average hourly heartrate. \\
@@
 \textbf{Feature Name} & \textbf{Feature description} \\
-\hline
+\midrule
 Breath Count & Number of breaths the person has made though an ACT  \\ 
  ~ & device during a treatment. A breath is counted if its maximum amplitude is \\ ~ & above the threshold of 8 mm H2O.\\
- \hline
+ \midrule
  Mean Breath Amplitude & Average breath amplitude (in mm H2O) during an ACT treatment. \\
@@
  Standard Deviation of Breath Amplitude & ~ \\
- \hline
+ \midrule
  Mean Breath Duration & Average duration (in seconds) of breath during an ACT treatment. \\ 
@@
  Standard Deviation of Breath Duration & ~ \\
- \hline
+ \midrule
  Break Count & Number of breaks (between breaths) in a treatment. \\
- \hline
+ \midrule
  Mean Break Duration & Average duration (in seconds) of breaks between breaths. \\ 
@@
  Standard Deviation of Break Duration & ~ \\
- \hline
+ \midrule
  Set Count & Number of sets in a treatment. Prescribed treatment may have up to 10 \\
  ~ & sets. Set is a group of breaths separated by huffing and coughing activity. \\
- \hline
+ \midrule
  Pressure Readings Count & Number of pressure reading registered by ACT device during a treatment. \\
  ~ & Used to filter out "test" device sessions that are not part of a treatment. \\
- \hline
+ \midrule
  Pressures Readings Completeness Score & Data completeness score for the day \\
@@
  ~ & score is 0. \\
- \hline
+ \midrule
  Treatment Duration & Treatment duration (in minutes). \\
- \hline
+ \midrule
  Treatment Daily Adherence Score & Number of completed treatments out of the prescribed ones (for the \\
  ~ & given day). \\
- \hline
+ \midrule
  Breath Count Daily Adherence Score & Number of registered breaths across all treatments for the given day \\
  ~ & out of the prescribed amount of breaths. \\ 
- \hline
+ \midrule
  Treatment Rolling Weekly Adherence Score & Number of completed treatments during the last 7 days (for a given date) \\
  ~ & out of the prescribed number of treatments for this period. \\
- \hline
+ \midrule
  Breath Count Rolling Weekly Adherence Score & Number of registered breaths across all treatments during the last 7 days \\
  ~ & (for a given date) out of the prescribed amount of breaths for this period. \\ 
- \hline
+ \midrule
  Treatment Weekly Adherence Score & Number of completed treatments during the last calendar week (for a given date) \\
  ~ & out of the prescribed number of treatments for this period. \\
- \hline
+ \midrule
  Breath Count Weekly Adherence Score & Number of registered breaths across all treatments during the last calendar week \\
  ~ & (for a given date) out of the prescribed amount of breaths for this period.\\
- \hline
+ \midrule
  Cumulative Adherence Score & Cumulative adherence score to date. \\
  ~ & How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount. \\
- \hline
+ \midrule
  Breathing Time Percentage & Time in a treatment spent on an exhalation, vs not.\\
- \hline
+ \midrule
  Skipped Days Count & Skipped days since the last treatment. \\
@@
     Cluster name & Cluster Size & Average Silhouette width \\
-    \hline
+    \midrule
     XS & 1099 & 0.95 \\
@@
     Cluster number & Cluster Size & Average Silhouette width \\
-    \hline
+    \midrule
     1 & 4996 & 0.12 \\
@@
     Cluster name & Cluster Size & Average Silhouette width \\
-    \hline
+    \midrule
     S & 1246 & 0.28 \\
@@
 
-Calinski-harabasz index for the heartrate clusters is 135.51 .
+Calinski-Harabasz index for the heartrate clusters is 135.51 .
 
@@
     Cluster number & Cluster Size & Average Silhouette width \\
-    \hline
+    \midrule
     1 & 4221 & -0.07 \\
@@
 \end{table}
+\section{Appendix : Cluster results and metrics for ACT treatments}
+
+The tables below show the metrics for the ACT treatment clusters.
+\subsection{K-Means Clustering}
+\subsubsection{K-Means Metrics and Number of Clusters}
+The table below shows comparison of Average Silhouette width and Calinski-Harabasz index different number of clusters
+
+\begin{table}[H]
+  \caption{Clusters Metrics Kmeans vs Number of Clusters}
+  \label{steps_metrics}
+  \centering
+  \begin{tabular}{ c|c|c}
+    \toprule
+    \cmidrule(r){1-2}
+    Number of Clusters & Average Silhouette width & Calinski-Harabasz index \\
+    \midrule
+    3 & 2292 & 0.17 \\
+    4 & 2208 & 0.18 \\
+    5 & 1996 & 0.15 \\
+    6 & 1830 & 0.15 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\subsubsection{Cluster interpretation}
+Clustering with number clusters k=4 has combination of good performance (Average Silhouette width and Calinski-Harabasz index) and positive feedback from clinical team on cluster level granularity and clusters interpretability.
+
+\begin{figure}
+  \centering
+  \caption{ACT clusters displayed on PCA and UMAP projections obtained using Kmeans clustering}
+  \includegraphics[scale=0.5]{fig_ACT_4_clusters.png}
+  \label{fig:figACT4clusters}
+\end{figure}
+
+\begin{table}[htb]
+  \caption{ACT Clusters interpretation}
+  \label{tab:act_interpret}
+  \centering
+  \begin{tabular}{ c|l}
+    \toprule
+
+  \textbf{Feature Name} & \textbf{Feature description} \\
+  \midrule
+    1 & Treatments with exceptionally big breath counts and biggest treatment adherence \\
+  \midrule
+    2 & Breath counts weekly adherence is on a lower side, treatments are short \\
+  \midrule
+    3 & Good breath counts and related breath weekly adherence, the breaths themselves have big standard  \\ 
+    ~ &  deviation of duration \\
+  \midrule
+    4 & Very good breath counts and related breath weekly adherence,  breaths are mostly short and powerful \\ 
+    ~ & (with biggest standard deviation for amplitude), shorter treatments  \\
+  \bottomrule
+    \end{tabular}
+\end{table}
 

### 2019-09-05 draft.tex
more additions to ACT appendix

@@
 
-We also assessed the clustering stability on different data partitions.
+We assessed the clustering stability on different data partitions.
 Random sampling 80~\% of the data and clustering using k-means provided very similar results with clustering on the full dataset: figure \ref{fig:ACTclusteringon80} shows that the shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index (2200 v 1770 respectively) were comparable.
@@
 \end{figure}
-
-*Not sure about this part. Maybe move to appendix?
-We also assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters.
-Figure \ref{fig:ACTclusteringAB} shows the same patient trajectory for clustering with feature selection A and clustering with feature selection B, which includes two additional features: breath amplitude standard deviation, and rolling weekly adherence score for breath count.
-It is notable that once computer gaming was introduced to ACT in both cases A and B the same patient moved from cluster with “expected” breath counts per ACT to a cluster characterized by much higher breath counts.
-This validates that there was a change over time in the patient’s ACT pattern that could be captured.
-The additional sdBreathAmplitude feature make results more explainable from a clinical point of view, as the original cluster for the patient in Clustering B was characterized by a centroid with the highest sdBreathAmplitude value.
-
-\begin{figure}[!htb]
-  \centering
-  \caption{Patient Trajectory on set A of features (without breath amplitude) and set B of features.}
-  \includegraphics[]{ACTclusteringAB.png}
-  \label{fig:ACTclusteringAB}
-\end{figure}
-
+We also assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters (see Appendix).
 \subsection{Clustering physical activity waveforms}
@@
     \cmidrule(r){1-2}
-    Number of Clusters & Average Silhouette width & Calinski-Harabasz index \\
+    Number of Clusters & Calinski-Harabasz index & Average Silhouette width \\
     \midrule
@@
 
+\subsection{Hierarchical Clustering}
+Figure \ref{fig:figACTdent} shows the Dendrogram of ACT data demonstrating major splits for number of clusterts k = 2,3,5 and 6
+
+\begin{figure}[htb]
+  \centering
+  \caption{ Hierarchical Clustering Dendrogram for ACT data}
+  \includegraphics[scale=0.5]{fig_ACT_HC_dentogram.png}
+  \label{fig:figACTdent}
+\end{figure}
+
+\subsubsection{Hierarchical Clustering Metrics and Number of Clusters}
+The table \ref{tab:act_hc_clusters} shows comparison of Average Silhouette width and Calinski-Harabasz index different number of clusters.
+
+\begin{table}[htb]
+  \caption{Clusters Metrics for Hierahical Clustering vs Number of Clusters}
+  \label{tab:act_hc_clusters}
+  \centering
+  \begin{tabular}{ c|c|c}
+    \toprule
+    \cmidrule(r){1-2}
+    Number of Clusters & Calinski-Harabasz index & Average Silhouette width \\
+    \midrule
+    3 & 1793 &  0.13\\
+    4 & 1643 &  0.13\\
+    5 & 1414 &  0.08\\
+    6 & 1289 &  0.06 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\begin{figure}[htb]
+  \centering
+  \caption{ACT clusters displayed on PCA and UMAP projections obtained using Hierarchical clustering}
+  \includegraphics[scale=0.5]{fig_ACT_HC_4_Clusters.png}
+  \label{fig:figHC4clusters}
+\end{figure}
+
+\section{Appendix : ACT Cluster validations via patient trajectories}
+We assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters.
+Figure \ref{fig:ACTclusteringAB} shows the same patient trajectory for clustering with feature selection A and clustering with feature selection B, which includes two additional features: breath amplitude standard deviation, and rolling weekly adherence score for breath count.
+It is notable that once computer gaming was introduced to ACT in both cases A and B the same patient moved from cluster with “expected” breath counts per ACT to a cluster characterized by much higher breath counts.
+This validates that there was a change over time in the patient’s ACT pattern that could be captured.
+The additional breath amplitude standard deviation feature (sdBreathAmplitude) make results more explainable from a clinical point of view, as the original cluster for the patient in Clustering B was characterized by a centroid with the highest sdBreathAmplitude value.
+
+\begin{figure}[htb]
+  \centering
+  \caption{Patient Trajectory on set A of features (without breath amplitude) and set B of features.}
+  \includegraphics[]{ACTclusteringAB.png}
+  \label{fig:ACTclusteringAB}
+\end{figure}
+
 \end{document}

### 2019-09-13 draft.tex
Saving progress as of Sept 12th.

@@
 Cystic Fibrosis (CF) is the most common life-limiting inherited disorder in the UK, affecting approximately 1 in 2500 babies born.
-It is a systemic genetic disorder that mainly affects the respiratory and digestive systems.
+It is a systemic genetic disorder that mainly affects the respiratory system.
 In the lungs excessive thick sticky mucus can cause cycles of infection, inflammation and lung damage leading to a deterioration in health.
@@
 
-To promote clearance of mucus from the lungs, daily physiotherapy is prescribed \cite{Daniels2017}.
-Airway clearance techniques (ACT) may include a series of  exhalations against resistance into hand-held ACT devices, followed by forced expiration techniques ('huffing') and coughing to improve mucociliary clearance and expel mucous; an exercise which is performed repeatedly \cite{Villanuevaj4574}.
-ACT is burdensome, with some CYPwCF reporting that ACT is the most burdensome part of having the disease \cite{JamesLindAlliance}.
+To promote clearance of mucus from the lungs, daily physiotherapy called airway clearance techniques (ACT) is prescribed \cite{Daniels2017}.
+Some CYPwCF reporting that ACT is the most burdensome part of having the disease \cite{JamesLindAlliance}.
 Physiotherapists also encourage children to have a physically active lifestyle, which has been shown to reduce the rate of decline in lung function \cite{Williams2014}.
@@
 
-The 18 month trial used an interrupted time series design, so that the effect of introducing and removing interventions such as gamification could be assessed [[*******UCL****** ref to details of trial design,is this sufficient?
-NREC:18-LO-1038 ].
+The 18 month trial used an interrupted time series design, so that the effect of introducing and removing interventions such as gamification could be assessed [NREC:18-LO-1038].
 Interventions included: introducing feedback to participants on their ACTs (number of breaths per treatment) via a computer tablet application, introducing games, and later removing these interventions.
 The data presented here is from the first 8x months of the trial.
-Participants were: [*******UCL****** exact numbers] 65x males and 80x females ranging in age from 6-16 years old, from 3 London hospitals, all of whom consented to participate in the study.
+Participants were: 71 females and 74 males ranging in age from 6-16 years old, from 3 London hospitals, all of whom consented to participate in the study.
 All data were completely de-identified for analysis.
@@
 As such, the data is unlabelled and we use unsupervised learning approach to achieve this.
-Unsupervised clustering is becoming an established technique for finding sub-types in physical activity data \cite{physical_activity_patterns_2017} and pressure data relating to respiratory disease [].
-
+Unsupervised clustering has been shown to be a useful technique for analysis of healthcare data, including finding novel sub-types in physical activity data \cite{physical_activity_patterns_2017} and sub-types in a variety of diseases including arthtritis \cite{Eng2019} and sepsis \cite{Seymour2019}.
 A high-scale data analysis and exploration pipeline was built in R in a secure analysis environment.
@@
 ACT treatments commonly involve around 10 sets of 10 exhalation breaths into an ACT device.
-Participants generally used 1 of 4 different ACT devices compatible with the pressure sensor; the Acapella, Aerobika, PariPEP or AstraPEP.
-Each device produced breaths with a slightly different overall shape, e.g.
-with small oscillations.
-Sometimes participants used different devices over different days.
-Each set helps to loosen mucus, and between sets, patients normally 'huff' and cough to expel it.
+Each set helps to loosen mucus, and between sets, patients normally 'huff' and cough to expel it \cite{Villanuevaj4574}.
 The treatments are usually performed one or more times per day depending on how well the person is or how much mucus they normally produce; participants did not have identical prescriptions.
@@
 A typical ACT waveform which has been pre-processed, is shown in figure \ref{fig:individual_breaths}.
-The defining characteristics of ACT treatments were summarised in a range of features (variables) including: breath count, breath duration, breath amplitude, and treatment duration.
-In order to derive these features, our breath identification algorithm recognises not only simple breaths, but complex breaths composed of multiple peaks, and having a variety of shapes & oscillations which are characteristic of the 3 different ACT devices used in the study.
-In addition to mean breath amplitude, we use amplitude standard deviation as a feature, in order to capture inconsistency or disorderliness in physiotherapy.
+The defining characteristics of ACT treatments were summarised into 27 features (variables) including: breath count, breath duration, breath amplitude, and treatment duration.
+In order to derive these features, our breath identification algorithm recognises not only simple breaths, but complex breaths composed of multiple peaks, and having a variety of shapes & oscillations which are characteristic of the 4 different ACT devices used in the study.
+We use amplitude standard deviation as a feature, in order to capture inconsistency or disorderliness in physiotherapy.
 We also identified how many discernible sets were apparent in the treatment.
@@
 
-\begin{figure}[htb]
-  \includegraphics[width=\textwidth]{individual_breaths.png}
+\begin{figure}
     \centering
+    \begin{minipage}{0.45\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{individual_breaths.png} % first figure itself
         \caption{A typical, pre-processed ACT waveform. This treatment is 20 seconds long and contains 7 exhalation breaths.}
         \label{fig:individual_breaths}
+    \end{minipage}\hfill
+    \begin{minipage}{0.45\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{HR.png} % second figure itself
+        \caption{One day showing heartrate data from the Fitbit with a sustained period of heartrate in MVPA.}
+         \label{fig:HR}
+    \end{minipage}
 \end{figure}
@@
 
-\begin{figure}[htb]
-  \includegraphics[width=\textwidth]{HR.png}
-  \centering
-   \caption{One day showing heartrate data with a sustained period of heartrate in MVPA.}
-  \label{fig:HR}
-\end{figure}
-
 Mindful of these data limitations, we extracted high-level patterns from the minute-averaged waveforms at a day level.
@@
 
-All features were verified by clinicians as being meaningful or potentially meaningful.
-We perform a correlation analysis on all features and removed features which were highly correlated.
-We use the Hopkins statistic [ref] to measure clustering tendency.
+We performed a correlation analysis on all features and removed features which were highly correlated, or were not verified by clinicians as being meaningful or potentially meaningful.
+We use the Hopkins statistic [ref] to measure clustering tendency (“factoextra” R package implementation implies that Hopkins statistic value close to 0 indicate the data is highly clusterable).
 Features with a Gaussian distribution were normalised to have mean~=~0 and standard deviation~=~1.
 In order to visualise the results, we used dimensionality reduction.
-We used the first two principle components from principal component analysis [ref], Uniform Manifold Approximation and Projection (UMAP) [ref], and t-Distributed Stochastic Neighbor Embedding (T-SNE [ref].
-We identified outliers using xxx (Olga) methods and removed less than xxx2~\%, with these cases being recognised as device malfunctions or abnormal behaviour.
+We used the first two principle components from principal component analysis (PCA) [ref], Uniform Manifold Approximation and Projection (UMAP) \cite{Mcinnes2018}, and t-Distributed Stochastic Neighbor Embedding (T-SNE) \cite{Maaten2008}.
+We identified outliers using Tukey method and removed less than xxx2~\%, with these cases being recognised as device malfunctions or abnormal behaviour.
 
-We employed the following clustering methods: k-means, Density-based spatial clustering of applications with noise (DBSCAN) [ref?] and hierarchical clustering.
+We employed the following clustering methods: k-means, Density-based spatial clustering of applications with noise and hierarchical clustering.
 
@@
 The features used for clustering are shown in table X below, while the full list of features is shown in appendix X.
+The Hopkins statistic for the data with the chosen features was 0.2069, suggesting  that data is has good clustering tendency.
 
-\subsubsection{Comparison of clustering techniques}
-We compared the three clustering techniques (see appendix) and found the best performance with k-means.
+\subsubsection{ACT K-means clustering results}
+The purpose of clustering into sub-types is to summarise how a participant is performing their ACT.
+We compared the K-means and Hierahical clustering techniques (see appendix) and found the best performance with k-means.
+We chose k=4 clusters to maximise both silhouette value (average 0.18), and the interpretability of the results.
+If k is too low, the sub-types are not granular enough, if k is too high it becomes challenging to remember the characteristics of each sub-type.The centroid values of the 4 clusters are shown in table X below.
 
@@
   \includegraphics[width=\textwidth]{featureTables.png}
-  \caption{To be replaced with a proper table.}
+  \caption{To be replaced with a proper table. Feature values for the cluster centroids.}
 \end{figure}
 
-\subsubsection{ACT K-means clustering results}
-The purpose of clustering into sub-types is to summarise how a participant is performing their ACT.
-We chose k=4 clusters to maximise both silhouette value (average 0.18), and the interpretability of the results.
-If k is too low, the sub-types are not granular enough, if k is too high it becomes challenging to remember the characteristics of each sub-type.
-
-The centroid values of the 4 clusters are shown in table X below. Based on these centroids, the four ACT sub-types, which we named alpha, beta, gamma and delta have the following interpretations:
+ Based on these centroids, the four ACT sub-types, which we named alpha, beta, gamma and delta have the following interpretations:
 
@@
 \end{itemize}
+
+The clusters capture variability of ACT treatments patient do at home, for example we identified clusters that are characterized by big adherence or breath with higher amplitude (see Appendix).
 
@@
 We assessed the clustering stability on different data partitions.
-Random sampling 80~\% of the data and clustering using k-means provided very similar results with clustering on the full dataset: figure \ref{fig:ACTclusteringon80} shows that the shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index (2200 v 1770 respectively) were comparable.
+Random sampling 80~\% of the data and clustering using k-means provided very similar results with clustering on the full dataset (see appendix), indicating stable clustering.
+We also assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters (see Appendix).
 
-\begin{figure}[!htb]
-  \centering
-  \caption{K-means clustering results on full dataset (left) and subset of data (right).}
-  \includegraphics[]{ACTclusteringOn80.png}
-  \label{fig:ACTclusteringon80}
-\end{figure}
-We also assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters (see Appendix).
 \subsection{Clustering physical activity waveforms}
@@
 Clustering of physical activity data was used to extract daily patterns from the steps and heart rate waveforms and characterise the physical activity levels and patterns of individual patients using a dataset containing 9760 individual days.
-The different clusters should extract more complex characteristics than the ones which could be obtained by simply combining some of the extracted features.
 
-\subsubsection{Steps clusters}
-
-The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, the mean of total step count per hour, the step cadence for the periods in MVPA and the number of hours in the day with 0~< ~steps~count~<~500.
-Features were selected as a result of feature selection, correlation and outlier analysis.
-Three sub-types were desired for the steps clusters covering small, medium and large amounts of steps activity throughout the day.
-Four clusters (as shown in figure \ref{fig:stepsClusters}) were chosen in this case to cover these three sub-types as well as a sub-type for the instances with missing data (XS cluster).
-
-\begin{figure}[htb]
-  \centering
-  \caption{Steps clusters displayed on the first two principle components.}
-  \includegraphics[]{stepsClusters.png}
-  \label{fig:stepsClusters}
-\end{figure}
-
-\begin{table}[H]
-  \caption{Steps Clusters Interpretation}
-  \label{steps_inter}
-  \centering
-  \begin{tabular}{ c|l }
-    \toprule
-    \cmidrule(r){1-2}
-    Cluster name & Cluster Interpretation \\
-    \midrule
-    XS & Waveforms with very little steps activity or high amount of missing data \\
-    \midrule
-    S & Moderate step activity, consistent cadence with few periods of high activity\\
-    \midrule
-    M & Consistent moderate to high step rate with a period of sustained high step activity \\
-    \midrule
-    L & High step rate with longer periods of sustained high step activity \\
-    \bottomrule
-    \end{tabular}
-\end{table}
+The two clustering methods used were K-means and hierarchical clustering with complete linkage.
+k-means clustering was chosen due to better performance on both the average silhouette width and the Calinski-Harabasz index as well as the resulting groupings on the dimensionality reduction plots (see Appendix for results).
+DBSCAN was also reviewed, but due to the nature of the data (continuous representation, flat structure) it was not used in the analysis.
 
@@
 The following were used as features for heart rate clustering: the number of minutes with heart rate higher than 120, coefficient of variation for heart rate values per day, number of time heart rate exceeds 100 and 120, and the median average hourly heart rate.
+The Hopkins statistic for the data with the chosen features was: 0.062, suggesting a high clustering tendency.
 Experiments were carried out with 3, 4 and 5 clusters and the choice of k was based on the cluster interpretability results.
 The sub-types found when k=5 were the best at identifying more complex patterns and grouping similar waveforms together.
-The five clusters found by K-means (as shown in figure \ref{fig:HRclusters} have the following interpretation:
+The five clusters found by K-means (as shown in figure \ref{fig:HRclusters}) have the following interpretation:
 
-\begin{figure}[!htb]
+\begin{figure}
     \centering
-  \includegraphics[]{HRclusters.png}
-  \caption{PCA of representation of the five heart rate clusters. }
+    \begin{minipage}{0.45\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{HRclusters.png} % first figure itself
+        \caption{PCA representation of the five heart rate clusters.}
         \label{fig:HRclusters}
+    \end{minipage}\hfill
+    \begin{minipage}{0.45\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{FitbitClusterValidation.png} % second figure itself
+        \caption{Steps clusters displayed on the first two principle components. Crosses are days with daily cadence above 30 which are all part of the L cluster.}
+ \label{fig:FitbitClusterValidation}
+    \end{minipage}
 \end{figure}
@@
 \begin{table}[H]
-  \caption{Heartrate Clusters Interpretation}
+  \caption{Heart rate Clusters Interpretation}
   \label{hr_inter}
   \centering
-  \begin{tabular}{ c|l }
+  \begin{tabular}{ c|d }
     \toprule
@@
 
+\subsubsection{Steps clusters}
+
+The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, the mean of total step count per hour, the step cadence for the periods in MVPA and the number of hours in the day with 0~< ~steps~count~<~500.
+Features were selected as a result of feature selection, correlation and outlier analysis.
+The Hopkins statistic for the data with the chosen features was: 0.13375, suggesting a high clustering tendency.
+Three sub-types were desired for the steps clusters covering small, medium and large amounts of steps activity throughout the day.
+Four clusters (as shown in figure \ref{fig:stepsClusters}) were chosen in this case to cover these three sub-types as well as a sub-type for the instances with missing data (XS cluster).
+
+\begin{table}[H]
+  \caption{Steps Clusters Interpretation}
+  \label{steps_inter}
+  \centering
+  \begin{tabular}{ c|d }
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster name & Cluster Interpretation \\
+    \midrule
+    XS & Waveforms with very little steps activity or high amount of missing data \\
+    \midrule
+    S & Moderate step activity, consistent cadence with few periods of high activity\\
+    \midrule
+    M & Consistent moderate to high step rate with a period of sustained high step activity \\
+    \midrule
+    L & High step rate with longer periods of sustained high step activity \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
 \subsubsection{Fitbit cluster validation }
@@
 Additionally, other features that were not included in the clustering analysis were used to validate the clusters and their interpretations.
-Certain expected conditions were checked against the cluster labels and plotted using the PCA graph such as figure \ref{fig:FitbitClusterValida} below.
+Certain expected conditions were checked against the cluster labels and plotted using the PCA graph such as figure \ref{fig:FitbitClusterValida}, which shows that all days with a cadence above 30, do indeed fall into the L cluster.
 Cluster boundaries for both steps and heartrate clusters were also utilised to validate the cluster properties and interpretation, these can be found in the Appendix.
-
-\begin{figure}[!htb]
-  \centering
-  \includegraphics[]{FitbitClusterValidation.png}
-  \caption{: Crosses are days with daily cadence above 30 which are all part of the L cluster.}
-  \label{fig:FitbitClusterValidation}
-\end{figure}
 
@@
 A minute is considered active if it has a step count > 100 and heartrate > 120 and is part of a sequence of at least $k$ consecutive minutes which satisfy both conditions.
+\\
 \midrule
-Heartrate Hourly max, min and median & Maximum. minimum and median of average hourly heartrate. \\
+Hourly Max/Min/Median Heartrate & Maximum, minimum and median of average hourly heartrate. \\
 \bottomrule
@@
 \cmidrule(r){1-2}
-\textbf{Feature Name} & \textbf{Feature description} \\
+\textbf{ACT features} & \textbf{Feature description} \\
 \midrule
-Breath Count & Number of breaths the person has made though an ACT  \\ 
- ~ & device during a treatment. A breath is counted if its maximum amplitude is \\ ~ & above the threshold of 8 mm H2O.\\
+Breath count & Number of breaths the person has made though an ACT device during a treatment. A breath is counted if its maximum amplitude is above the threshold of 8 mm H2O.\\
  \midrule
- Mean Breath Amplitude & Average breath amplitude (in mm H2O) during an ACT treatment. \\
- Minimum Breath Amplitude & Similarly minimum, maximum and standard deviation of breath amplitude. \\
- Maximum Breath Amplitude & ~ \\
- Standard Deviation of Breath Amplitude & ~ \\
+ Mean/Min/Max/Standard Deviation of breath amplitude & Statistics of breath amplitude (in mm H2O) during an ACT treatment. \\
   \midrule
- Mean Breath Duration & Average duration (in seconds) of breath during an ACT treatment. \\ 
- Minimum Breath Duration &  Similarly minimum, maximum and standard deviation of \\ 
- Maximum Break Duration & breath duration in a treatment. \\
- Standard Deviation of Breath Duration & ~ \\
+ Mean/Min/Max/Standard Deviation of breath duration & Statistics of breath duration (in seconds) during an ACT treatment. \\
  \midrule
- Break Count & Number of breaks (between breaths) in a treatment. \\
+ Break count & Number of breaks (between breaths) in a treatment. \\
  \midrule
- Mean Break Duration & Average duration (in seconds) of breaks between breaths. \\ 
- Minimum Break Duration & Similarly minimum, maximum and standard deviation of \\ 
- Maximum Break Duration & breaks duration in a treatment. \\
- Standard Deviation of Break Duration & ~ \\
+ Mean/Min/Max/Standard Deviation of breath duration & Statistics of duration (in seconds) of breaks between breaths in an ACT treatment. \\
  \midrule
- Set Count & Number of sets in a treatment. Prescribed treatment may have up to 10 \\
- ~ & sets. Set is a group of breaths separated by huffing and coughing activity. \\
+ Set count & Number of sets in a treatment. Prescribed treatment may have up to 10 sets. Set is a group of breaths separated by huffing and coughing activity. \\
  \midrule
- Pressure Readings Count & Number of pressure reading registered by ACT device during a treatment. \\
- ~ & Used to filter out "test" device sessions that are not part of a treatment. \\
+ Pressure readings count & Number of pressure reading registered by ACT device during a treatment. Used to filter out "test" device sessions that are not part of a treatment. \\
  \midrule
- Pressures Readings Completeness Score & Data completeness score for the day \\
- ~ & If a treatment has less than 150 pressure readings it's completeness \\
- ~ & score is 0. \\
+ Pressures readings completeness score & Data completeness score for the day If a treatment has less than 150 pressure readings it's completeness score is 0. \\
  \midrule
- Treatment Duration & Treatment duration (in minutes). \\
+ Treatment duration & Treatment duration (in minutes). \\
  \midrule
- Treatment Daily Adherence Score & Number of completed treatments out of the prescribed ones (for the \\
- ~ & given day). \\
+ Treatment daily adherence score & Number of completed treatments out of the prescribed ones (for the given day). \\
  \midrule
- Breath Count Daily Adherence Score & Number of registered breaths across all treatments for the given day \\
- ~ & out of the prescribed amount of breaths. \\ 
+ Breath count daily adherence score & Number of registered breaths across all treatments for the given day out of the prescribed amount of breaths. \\ 
  \midrule
- Treatment Rolling Weekly Adherence Score & Number of completed treatments during the last 7 days (for a given date) \\
- ~ & out of the prescribed number of treatments for this period. \\
+ Treatment rolling weekly adherence score & Number of completed treatments during the last 7 days (for a given date) out of the prescribed number of treatments for this period. \\
  \midrule
- Breath Count Rolling Weekly Adherence Score & Number of registered breaths across all treatments during the last 7 days \\
- ~ & (for a given date) out of the prescribed amount of breaths for this period. \\ 
+ Breath count rolling weekly adherence score & Number of registered breaths across all treatments during the last 7 days (for a given date) out of the prescribed amount of breaths for this period. \\ 
  \midrule
- Treatment Weekly Adherence Score & Number of completed treatments during the last calendar week (for a given date) \\
- ~ & out of the prescribed number of treatments for this period. \\
+ Treatment weekly adherence score & Number of completed treatments during the last calendar week (for a given date) out of the prescribed number of treatments for this period. \\
  \midrule
- Breath Count Weekly Adherence Score & Number of registered breaths across all treatments during the last calendar week \\
- ~ & (for a given date) out of the prescribed amount of breaths for this period.\\
+ Breath count weekly adherence score & Number of registered breaths across all treatments during the last calendar week(for a given date) out of the prescribed amount of breaths for this period.\\
  \midrule
- Cumulative Adherence Score & Cumulative adherence score to date. \\
- ~ & How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount. \\
+ Cumulative adherence score & Cumulative adherence score to date. How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount. \\
  \midrule
- Breathing Time Percentage & Time in a treatment spent on an exhalation, vs not.\\
+ Breathing time percentage & Percentage of time in a treatment spent on an exhalation \\
  \midrule
- Skipped Days Count & Skipped days since the last treatment. \\
+ Skipped days count & Skipped days since the last treatment. \\
  \bottomrule
@@
 
-The tables below show the metrics for the clusters identified for the physical activity data.
-
 \subsection{Steps clusters}
 
-\subsubsection{K-Means Clustering}
+\subsubsection{K-means Clustering}
 
@@
 
-The clustering results using Hierarchical clustering with complete linkage on steps data can be seen in Figure \ref{fig:stepsClustersHC}.
+The clustering results using Hierarchical clustering with complete linkage on steps data displayed using the first two principal components can be seen in Figure \ref{fig:stepsClustersHC}.
 
@@
   \centering
-  \begin{tabular}{ |c|c|c|}
+  \begin{tabular}{ c|c|c}
     \toprule
@@
 
-The clustering results using Hierarchical clustering with complete linkage for the heartrate data can be seen in Figure \ref{fig:hrClustersHC}.
+The clustering results using Hierarchical clustering with complete linkage for the heartrate data displayed using the first two principal component can be seen in Figure \ref{fig:hrClustersHC}.
 
@@
   \centering
-  \begin{tabular}{ c|l}
+  \begin{tabular}{ c|d}
     \toprule
 
-  \textbf{Feature Name} & \textbf{Feature description} \\
+  \textbf{Cluster} & \textbf{Cluster Interpretation} \\
   \midrule
@@
   \midrule
-    3 & Good breath counts and related breath weekly adherence, the breaths themselves have big standard  \\ 
-    ~ &  deviation of duration \\
+    3 & Good breath counts and related breath weekly adherence, the breaths themselves have big standard deviation of duration \\
   \midrule
-    4 & Very good breath counts and related breath weekly adherence,  breaths are mostly short and powerful \\ 
-    ~ & (with biggest standard deviation for amplitude), shorter treatments  \\
+    4 & Very good breath counts and related breath weekly adherence,  breaths are mostly short and powerful (with biggest standard deviation for amplitude), shorter treatments  \\
   \bottomrule
@@
 
+\section{Appendix : ACT clustering with different data partitions}
+We assessed the clustering stability on different data partitions.
+Random sampling 80~\% of the data and clustering using k-means provided very similar results with clustering on the full dataset: figure \ref{fig:ACTclusteringon80} shows that the shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index (2200 v 1770 respectively) were comparable.
+
+\begin{figure}[!htb]
+  \centering
+  \caption{K-means clustering results on full dataset (left) and subset of data (right).}
+  \includegraphics[]{ACTclusteringOn80.png}
+  \label{fig:ACTclusteringon80}
+\end{figure}
+
 \end{document}

### 2019-09-13 final.tex
Saving progress as of Sept 12th.

@@
 \usepackage{graphicx}
+\usepackage{float}
+\usepackage{longtable}
+\usepackage{multirow}
+\usepackage{array}
+\usepackage{tabularx}
+\usepackage{diagbox}
+\newcolumntype{n}{>{\arraybackslash}m{5.3cm}}
+\newcolumntype{d}{>{\arraybackslash}m{8cm}}
 
-\title{A data-driven study of cystic fibrosis treatment: identifying sub-types of physiotherapy style and physical activity using unsupervised clustering}
+\title{A data-driven study of physiotherapy for cystic fibrosis: identifying sub-types of airway clearance techniques and physical activity patterns using unsupervised clustering}
 
@@
 \begin{abstract}
-  Abstract here
+  Children and young people with cystic fibrosis (CYPwCF) are prescribed physiotherapy including daily airway clearance techniques (ACT), and being physically active as part of their treatment.
+Little is known about how CYPwCF follow this advice in their everyday lives.
+Physiotherapy has evolved over the decades without the clarity of evidence that would accrue from detailed data.
+Such data are now being collected as part of Project [Anonymous] which is the largest CF physiotherapy trial in the UK to date.
+The data we analyse come firstly from novel pressure sensors attached to existing ACT devices which measure breathing patterns during normal treatments, and secondly from Fitbit activity trackers which monitor physical activity patterns, from 139 participants over 9 months.
+In close collaboration with clinical experts, we quantified and characterised this unique ACT and activity data by creating custom features from the waveforms.
+Using unsupervised learning techniques, we identified meaningful sub-types, with k-means clustering achieving the best performance.
+We found 4 clusters of ACT data which were sensitive enough to track interventions in ACT routines.
+Within physical activity data we identified 4 clusters related to daily footsteps and 5 clusters related to heart rate data.
 \end{abstract}
 
+\section{Problem overview}
+
+Cystic Fibrosis (CF) is the most common life-limiting inherited disorder in the UK.
+It is a genetic disorder that mainly affects the respiratory system.
+In the lungs, excessive thick sticky mucus can cause lung damage leading to a deterioration in health.
+Despite improvements in care, CF remains progressive and incurable.
+
+To promote clearance of mucus from the lungs, daily physiotherapy called airway clearance techniques (ACT) is prescribed \cite{Daniels2017}.
+ACT commonly involve around 10 sets of 10 exhalation breaths into an ACT device, once to twice per day.
+Some CYPwCF report that ACT is the most burdensome part of having the disease \cite{JamesLindAlliance}.
+Physiotherapists also encourage children to have a physically active lifestyle, which has been shown to reduce the rate of decline in lung function \cite{Williams2014}.
+Adherence to physiotherapy has traditionally been assessed via questionnaire or self report, both of which are known to be inaccurate.
+More objective measures of adherence have not been widely available, and the evolution of CF physiotherapy has not been informed by this important evidence \cite{Modi2006}.
+
+Previously, bespoke pressure sensors were developed which attach to existing ACT devices in order to facilitate remote monitoring of ACT undertaken routinely at home (recording pressure values in time-stamped millibars, sampling frequency 10~Hz) \cite{Zhang}.
+Previously, bespoke games were also developed which could be driven by pressure changes from breathing through the sensor.
+These are now being used in a study of 145 CYPwCF \cite{HelenDouglas2018} to quantify routine ACTs (using the pressure sensors) and physical activity patterns (using Fitbit activity trackers), and to study the effects of gamification.
+This is the largest CF physiotherapy trial in the UK, collecting completely novel daily data about how children follow physiotherapy advice in their everyday lives.
+The broader goals of the study are to link physiotherapy adherence patterns with clinical outcomes (like lung function) in order to optimise and personalise treatment, which could reduce treatment burden and improve patient outcomes.
+In this paper we describe the challenge of i) quantifying ACT pressure sensor data and Fitbit activity data by creating custom features of the waveforms (featurising), and ii) discovering sub-types of ACT and physical activity patterns using clustering techniques.
+These daily sub-types can be tracked over time, as interventions such as gaming are introduced.
+
+\section{Methods}
+
+\subsection{Study design and analysis approach}
+
+The 18 month study \cite{HelenDouglas2018} includes the following interventions: introducing feedback to participants on their ACTs (number of breaths per treatment) via a computer tablet application, introducing games, and later removing these interventions.
+The data presented here is from the first 9 months of the trial with 139 participants enrolled at staggered times.
+Participants ranged in age from 6-16 years old, came from 3 London hospitals, all consented to participate in the study, and all data were completely de-identified for analysis.
+
+Our analysis seeks to find patterns in physiotherapy data, and as such, the data is unlabelled and we use an unsupervised learning approach.
+Unsupervised clustering has been shown to be a useful technique for analysis of healthcare data, including finding novel sub-types in physical activity data from wearables \cite{physical_activity_patterns_2017} and sub-types in a variety of diseases including arthtritis \cite{Eng2019} and sepsis \cite{Seymour2019}.
+A high-scale data analysis and exploration pipeline was built in R in a secure analysis environment.
+
+\subsection{Collaboration, trust and interpretability}
+
+Many applications of machine learning to medical data fail to explain how (if at all) the expertise of clinicians has been included \cite{Alaa2018}.
+To address the issues that arise from a lack of collaboration \cite{Vayena2018}, \cite{Challen231}, \cite{Char2018}, \cite{Ahmad2018} our methodology involved close collaboration between data scientists and CF experts, making every data decision jointly.
+We used a novel approach of involving at least one clinical expert in engineering “stand-ups” every day over the course of building the data analysis pipeline (7 months).
+We only considered features that were meaningful to clinicians, and we balanced traditional statistical evaluation metrics, with what was interpretable to clinicians.
+
+\section{Experiments and results}
+
+\subsection{Quantifying physiotherapy data and creating clinical features}
+
+\subsubsection{Featurising ACT data}
+
+Several pre-processing steps were performed on the raw pressure data which were a function of the custom sensor used, including assembling data packets and removing duplicate values.
+The sensor also introduced significant non-linear baseline drift, which was removed with a sparsity-based de-noising approach with an asymmetry factor, called BEADS \cite{Ning2014}.
+An example of pre-processing raw pressure data is shown in figure \ref{fig: Waveforms}.
+
+\begin{figure}[htb]
+  \includegraphics[width=\textwidth]{Waveforms.png}
+  \centering
+   \caption{Top left: raw ACT data in two packets, with baseline drift.
+Top right: Data packets assembled into a treatment and baseline drift removed.
+This treatment shows 11 sets of 10 exhalations (the positive peaks), lasting 33 minutes.
+Bottom left: A typical, pre-processed ACT waveform.
+This treatment is 20 seconds long and contains 7 exhalation breaths.
+Bottom right: One day showing heartrate data from the Fitbit with a sustained period of heartrate in MVPA.}
+  \label{fig: Waveforms}
+\end{figure}
+
+In total, we recorded 24,434 sensor readings.
+This included data from every time the ACT sensor was turned on, including for testing and demonstration.
+Once we removed readings shorter than 15 seconds, treatments with no breath exhalations, and situations with sensor misuse/malfunction, we were left with 14,689 ACT treatments for analysis.
+
+A typical ACT waveform which has been pre-processed, is shown in figure \ref{fig:individual_breaths}.
+The defining characteristics of ACT treatments were summarised features (variables) such as number of breaths in a treatment.
+In order to derive these features, we used peak detection to not only identify simple breaths as shown in figure \ref{fig: Waveforms}, but complex breaths composed of multiple peaks, and having a variety of shapes & oscillations which are characteristic of the 4 different ACT devices used in the study.
+%\begin{figure}
+%    \centering
+%    \begin{minipage}{0.48\textwidth}
+%        \centering
+%        \includegraphics[width=0.9\textw%idth]{individual_breaths.jpg} % first %figure itself
+%        \caption{A typical, %pre-processed ACT waveform. This %treatment is 20 seconds long and %contains 7 exhalation breaths.}
+%        \label{fig:individual_breaths}
+%    \end{minipage}\hfill
+%    \begin{minipage}{0.48\textwidth}
+%        \centering
+%        \includegraphics[width=0.88\text%width]{HR.jpg} % second figure itself
+%        \caption{One day showing %heartrate data from the Fitbit with a %sustained period of heartrate in MVPA.}
+%         \label{fig:HR}
+%    \end{minipage}
+%\end{figure}
+
+We use breath amplitude standard deviation as a feature, in order to capture inconsistency or disorderliness in physiotherapy.
+Most features are on a per-treatment level.
+It is common for prescribed treatments to be skipped, and days with no treatments is an important signal itself.
+Several features were derived which captured missing data: treatment adherence to date, skipped days since last treatment, rolling adherence over 7 days.
+These features are on a per-day level, as participants are prescribed 1 or 2 treatments per day.
+We created a total of 27 features (Appendix \ref{appendix:actfeatures}) to describe ACT data.
+
+\subsubsection{Featurising Fitbit data}
+
+The Fitbit activity tracker (Fitbit AltaHR) provided each participant’s step count per minute.
+While real-time heartbeat data would allow for a detailed analysis of heart rate waveforms, the Fitbit only provides average heart rate, with a variable sampling frequency ranging from 3 to 12 values per minute.
+To achieve consistent frequency and comparable values, the heart rate values were averaged for each minute.
+An example of heart rate data is shown in figure \ref{fig: Waveforms}.
+
+We extracted high-level patterns from the minute-averaged waveforms at a day level.
+The heart rate and steps data sets were featurised using counts per day, such as number of minutes in moderate to vigorous intensity physical activities (MVPA) (heart rate above 120 bpm) per day.
+To account for missing data during times that the Fitbit was not worn, the wear percent of the Fitbit was calculated from the valid heart rate values.
+This measure was used to select only days with a substantial portion of data for the day (80~\% of wear time from 6~am to 6~pm), resulting in a dataset containing 9760 days used for analysis.
+We created a total of 80 features (Appendix \ref{appendix:fitbitfeatures}) to describe Fitbit data.
+
+\subsection{Clustering physiotherapy data to find sub-types}
+
+We removed features which were highly correlated, and which were not verified by clinicians as being meaningful.
+We normalised all features and identified outliers using Tukey's method and removed less than 5~\%, with these cases being manually validated as device malfunctions or abnormal behaviour.
+
+We performed an initial assessment of clustering methods:  k-means, density-based spatial clustering of applications with noise (DBSCAN) and hierarchical clustering (HC) (see Appendices \ref{appendix:fitbitclusters} and \ref{appendix:actclusters}).
+For both ACT and Fitbit data, we found K-means clustering had the best performance in terms of silhouette width, Calinski-Harabasz index as well as having the most interpretable, clinically intuitive clusters.
+We selected k-means which we use for the rest of the analyses.
+
+\subsubsection{Finding sub-types in ACT data}
+
+Some features were excluded from clustering, but provided interesting physiotherapy insights: the number of sets in a treatment were expected to be influential in clustering, as a regular pattern of sets would show close adherence to physiotherapy advice.
+However we found that only 15\% of treatments had discernible sets, so this feature could not be used.
+The 7xx features used for clustering are shown in table X below, while the full list of features is shown in appendix \ref{appendix:actfeatures}.
+The Hopkins statistic for the data with the chosen features was 0.198, suggesting a high clustering tendency.
+
+The purpose of clustering into sub-types is to summarise how a participant is performing their ACT.
+We found the best clustering performance with k=4; the with centroid values are shown in table X below.
+
+\begin{table}[H]
+  \caption{ACT sub-types: clusters and centroid values. *: Fractional score calculated over a rolling 7 day period, all other features calculated at a treatment level.}
+  \label{ACT_centroids}
+  \centering
+  \begin{tabularx}{17cm}{ X|X|X|X|X|X|X|X|X}
+    \toprule
+    \cmidrule(r){1-2}
+    ACT sub-type & 
+    Number of breaths & Mean breath duration (s) & SD breath duration (s) & Mean breath amplitude (mBar) &  SD breath amplitude (mBar) & Treatment duration (m) & 7-day treatment adherence* & 7-day breath adherence*  \\
+    \midrule
+    Alpha & 83.1645 & 1.1324091 & 0.9544094 & 19.13988 & 4.317706 & 11.31548 & 0.6117628 & 0.7350068\\
+    Beta & 123.4091 & 0.8339444& 0.9706392 & 36.11484 & 9.864017 & 12.85468 & 0.7966647 & 1.2348730\\
+    Gamma & 106.7632&     2.2134100 & 3.2749929 & 23.91718 & 6.191824 &  22.33150 &  0.7410448 &  1.1968865   \\
+    Delta & 203.8484  &    1.1585528 & 1.2114907 &   21.55130 & 5.471787 & 26.88816 &  0.8643425 & 1.9809176 \\
+    \bottomrule
+    \end{tabularx}
+\end{table}
+
+The clusters identify sub-types in ACT data, for example the Alpha sub-type is characterised by near ideal (20 mBar) breath amplitude, low standard deviations (SD) indicating consistency in technique, and lower adherence, further interpretations in appendix \ref{appendix:actKmeansinterpretation}.
+These qualitative findings are consistent across two dimensionality reduction techniques, PCA and UMAP \cite{Mcinnes2018}.
+
+\subsubsection{ACT cluster validation }
+
+Although the study is not complete, early results indicate that clusters are sensitive enough to track changes in some participants' adherence through different study interventions, as shown in figure ~\ref{fig:act_res}.
+We found that clustering was stable based on different data partitions, in terms of metrics such as silhouettes, and in terms of consistency of participant's cluster membership (see appendix \ref{appendix:actdatapartitions} and appendix \ref{appendix:actpatienttrajectories}).
+
+\begin{figure}[!htb]
+  \centering
+  \includegraphics[]{fig_ACT_clust}
+  \caption{the four ACT clusters overlaid with a single participant (black), showing that the participant’s cluster membership changes over time.
+Clustering results are visualised in 2-dimensions using PCA (left) and UMAP (right).
+The time points are before ACT feedback (number of registered breaths) is introduced (top), after ACT feedback is introduced (middle), and once computer gaming is introduced to ACT routine (bottom).
+Each point represents one treatment of one participant.
+}
+  \label{fig:act_res}
+\end{figure}
+
+\subsubsection{Finding sub-types in Fitbit data}
+
+The purpose of clustering into sub-types is to summarise each participant's steps and heart rate waveforms, using a dataset containing 9760 days.
+
+\subsubsubsection{Heart rate clusters}
+
+The following were used as features for heart rate clustering: the number of minutes with heart rate higher than 120, coefficient of variation for heart rate values per day, number of time heart rate exceeds 100 and 120, and the median average hourly heart rate.
+The Hopkins statistic for the data with the chosen features was: 0.062, suggesting a high clustering tendency.
+We found the best clustering performance with k=5, with clusters shown in figure \ref{fig:HRclusters}.
+The heart rate cluster centroid values are shown in Figure \ref{fig:hrBoundaries}.
+The clusters identify sub-types in heart rate data, for example the M1 sub-type is characterised Waveforms with high variability and periods with spikes above 120 bpm (see appendix \ref{appendix:hrclusters} for more details).
+
+\begin{figure}
+    \centering
+    \begin{minipage}{0.45\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{HRclusters.png} % first figure itself
+        \caption{PCA representation of the five heart rate clusters.}
+        \label{fig:HRclusters}
+    \end{minipage}\hfill
+    \begin{minipage}{0.45\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{FitbitClusterValidation.jpg} % second figure itself
+        \caption{Steps clusters displayed on the first two principle components. Crosses are days with daily cadence above 30 which are all part of the L cluster.}
+ \label{fig:FitbitClusterValidation}
+    \end{minipage}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.5]{hr_cluster_boundaries.png}
+  \caption{Heart rate Cluster Boundaries}
+  \label{fig:hrBoundaries}
+\end{figure}
+
+\subsubsubsection{Steps clusters}
+
+The following features were used for clustering steps: number of minutes with step count higher than 100 for at least 5 consecutive minutes, mean step count per hour, step cadence for periods in MVPA, and number of hours in the day with 0~< ~step~count~<~500.
+The Hopkins statistic for the data with the chosen features was: 0.13375, suggesting a high clustering tendency.
+Four clusters were chosen, including one for missing data with the cluster boundaries shown in Figure \ref{fig:stepsBoundaries} (see appendix \ref{appendix:stepsclusters}).
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.5]{steps_cluster_boundaries.png}
+  \caption{Steps Cluster Boundaries}
+  \label{fig:stepsBoundaries}
+\end{figure}
+
+\subsubsection{Fitbit cluster validation }
+
+Steps and heart rate clusters were validated with clinical experts and found to be meaningful and useful ways to categorise the waveforms.
+Individual waveforms such as figure \ref{fig:HR} were manually inspected together with the cluster boundaries to ensure consistency across cluster interpretations and the actual cluster characteristics.
+Certain expected conditions were checked against the cluster labels and plotted using the PCA graph such as figure \ref{fig:FitbitClusterValidation}, which shows that all days with a cadence above 30, do indeed fall into the L cluster.
+
+\section{Discussion and conclusion}
+
+Project [Anonymised] involves collecting novel data about how CF physiotherapy advice on ACTs and physical activity is adhered to in everyday life, information which has never been available to physiotherapists.
+In close collaboration with CF experts, we developed methods to interrogate and make sense of this data.
+Project [Anonymised] is an ongoing study, with participation lasting 18 months.
+8 months into the trial we do not draw any conclusions, as we do not have an equal number of days of data from patients recruited over 8 months.
+Instead, we present a framework for analysis, which will empower clinicians to understand how physiotherapy advice is actually adhered to, and interpret the impact of interventions such as gamification when the trial is complete.
+Our framework involves processing the ACT and Fitbit data, deriving clinically meaningful features from it which characterise this novel data, and then using unsupervised clustering to find sub-types of this data.
+The clusters provide a useful way of tracking participants' trajectories across sub-types over time.
+In future the ACT and Fitbit features can be combined with each other, and with clinical data, to understand what type of physiotherapy leads to better clinical outcomes.
+The analysis of this data could help lead to more personalised treatment for CF, which could reduce treatment burden and improve health outcomes.
+
+\bibliographystyle{abbrv}
+% include your .bib file
+\bibliography{fizzyo}
+
+\newpage
+\section*{Appendices}
+\appendix
+
+\section{Table of physical activity data features}
+\label{appendix:fitbitfeatures}
+
+\begin{longtable}{  n | d }
+\toprule
+\cmidrule(r){1-2}
+\textbf{Fitbit features} & \textbf{Feature description} \\
+\midrule
+\endfirsthead
+Minutes wear & Number of minutes in a day with heartrate value present \\
+\midrule
+Percentage of wear & $\frac{\textit{Minutes wear}}{24\cdot60}\cdot100\%$\\
+\midrule
+Wear during sleep & Indicator value: 1,  worn during sleep or 0, otherwise \\
+\midrule
+Active hours & Number of hours with a step count > 500  \\
+\midrule
+Low active hours & Number of hours with a step count < 500  \\
+\midrule
+Mean hourly steps & Average of hourly step count \\
+\midrule
+Step count & Total step count per day \\
+\midrule
+Normalised step count & Rate of steps per minute when fitbit is worn $\frac{\textit{Step Count}}{\textit{Minutes Worn}}$ \\
+\midrule
+ Step active minutes $k$, for $k\in \{0,2,5,10,20\}$ & Number of active minutes in a day.
+A minute is considered active if it has a step count > 100 and is part of a sequence of at least $k$ consecutive minutes with step count > 100.
+Note that for $k=0$ there is no restriction on the number of consecutive minutes\\
+\midrule
+Step active minutes during waking period & Number of minutes with step > 100 from 6am to midnight \\
+\midrule
+Step coefficient of variation & Coefficient of variation of minute step counts $CoV = \frac{\sigma (\textit{Step count})}{\mu (\textit{Step count})} \cdot 100$ \\
+ \midrule
+ Normalised step count for periods of : &  Rate of steps per minute when the heartrate is: \\
+ - moderate activity & - between 120 and 140. \\ 
+ - vigorous activity & - above 140. \\
+ - moderate and vigorous activity & - above 120. \\
+ - non-moderate and non-vigorous activity & - below 120.\\
+ \midrule
+ Resting heartrate proxy & Mean of the 5 minutes with the lowest heartrate average \\
+ \midrule
+Heartrate active minutes $k$, for $k\in \{0,2,5,10,20\}$ & Number of active minutes in a day.
+A minute is considered active if it has heartrate > 120 and is part of a sequence of at least $k$ consecutive minutes with heartrate > 120.\\
+\midrule
+Heartrate active minute during waking period & Number of minutes with heartrate > 120 from 6am to midnight \\
+\midrule
+Heartrate coefficient of variation & Coefficient of variation for heartrate minute averages $CoV = \frac{\sigma (\textit{Heartrate})}{\mu (\textit{Heartrate})} \cdot 100$ \\
+ \midrule
+ Number of switches above threshold $t$, for $t\in \{100,120,140\}$ & Number of times the heartrate goes above the threshold $t$ from below this threshold.\\ 
+ \midrule
+ Step and heartrate active minutes $k$, for $k\in \{0,2,5,10,20\}$ & Number of active minutes in a day.
+A minute is considered active if it has a step count > 100 and heartrate > 120 and is part of a sequence of at least $k$ consecutive minutes which satisfy both conditions.
+\\
+\midrule
+Hourly Max/Min/Median heart rate & Maximum, minimum and median of average hourly heartrate. \\
+\bottomrule
+\end{longtable}
+
+\section{Table of ACT treatment features}
+\label{appendix:actfeatures}
+
+\begin{longtable}{ n|d}
+\toprule
+\cmidrule(r){1-2}
+\textbf{ACT features} & \textbf{Feature description} \\
+\midrule
+Breath count & Number of breaths the person has made though an ACT device during a treatment. A breath is counted if its maximum amplitude is above the threshold of 8 mm H2O.\\
+ \midrule
+ Mean/Min/Max/Standard Deviation of breath amplitude & Statistics of breath amplitude (in mm H2O) during an ACT treatment. \\
+  \midrule
+ Mean/Min/Max/Standard Deviation of breath duration & Statistics of breath duration (in seconds) during an ACT treatment. \\
+ \midrule
+ Break count & Number of breaks (between breaths) in a treatment. \\
+ \midrule
+ Mean/Min/Max/Standard Deviation of breath duration & Statistics of duration (in seconds) of breaks between breaths in an ACT treatment. \\
+ \midrule
+ Set count & Number of sets in a treatment. Prescribed treatment may have up to 10 sets. Set is a group of breaths separated by huffing and coughing activity. \\
+ \midrule
+ Pressure readings count & Number of pressure reading registered by ACT device during a treatment. Used to filter out "test" device sessions that are not part of a treatment. \\
+ \midrule
+ Pressures readings completeness score & Data completeness score for the day If a treatment has less than 150 pressure readings it's completeness score is 0. \\
+ \midrule
+ Treatment duration & Treatment duration (in minutes). \\
+ \midrule
+ Treatment daily adherence score & Number of completed treatments out of the prescribed ones (for the given day). \\
+ \midrule
+ Breath count daily adherence score & Number of registered breaths across all treatments for the given day out of the prescribed amount of breaths. \\ 
+ \midrule
+ Treatment rolling weekly adherence score & Number of completed treatments during the last 7 days (for a given date) out of the prescribed number of treatments for this period. \\
+ \midrule
+ Breath count rolling weekly adherence score & Number of registered breaths across all treatments during the last 7 days (for a given date) out of the prescribed amount of breaths for this period. \\ 
+ \midrule
+ Treatment weekly adherence score & Number of completed treatments during the last calendar week (for a given date) out of the prescribed number of treatments for this period. \\
+ \midrule
+ Breath count weekly adherence score & Number of registered breaths across all treatments during the last calendar week(for a given date) out of the prescribed amount of breaths for this period.\\
+ \midrule
+ Cumulative adherence score & Cumulative adherence score to date. How many in total treatments patient done over a period (since the beginning of the study out the prescribed amount. \\
+ \midrule
+ Breathing time percentage & Percentage of time in a treatment spent on an exhalation \\
+ \midrule
+ Skipped days count & Skipped days since the last treatment. \\
+ \bottomrule
+\end{longtable}
+
+\section{Cluster results and metrics for physical activity data}
+\label{appendix:fitbitclusters}
+\subsection{Steps clusters}
+\label{appendix:stepsclusters}
+\subsubsection{K-means Clustering}
+\label{appendix:stepsKmeans}
+
+\paragraph{Clusters Interpretation}
+Clusters interpretation for the four steps clusters can be seen in Table \ref{steps_inter}.
+
+\begin{table}[H]
+  \caption{Steps Clusters Interpretation}
+  \label{steps_inter}
+  \centering
+  \begin{tabular}{ c|d }
+    \toprule
+    \cmidrule(r){1-2}
+    \textbf{Cluster name} & \textbf{Cluster Interpretation} \\
+    \midrule
+    XS & Waveforms with very little steps activity or high amount of missing data \\
+    \midrule
+    S & Moderate step activity, consistent cadence with few periods of high activity\\
+    \midrule
+    M & Consistent moderate to high step rate with a period of sustained high step activity \\
+    \midrule
+    L & High step rate with longer periods of sustained high step activity \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\paragraph{Clusters Boundaries}
+Clusters boundaries  for the four steps clusters can be seen in Table \ref{steps_boundaries}.
+
+\begin{table}[H]
+  \caption{Steps clusters boundaries - mean, median, min, max}
+  \label{steps_boundaries}
+  \centering
+  \begin{tabular}{p{4cm}*{5}{|p{1cm}}}
+    \toprule
+    \multicolumn{2}{c|}{\backslashbox{Boundaries}{Cluster Name}} & \textbf{XS} & \textbf{S} & \textbf{M} & \textbf{L} \\
+    \midrule
+    \multirow{4}{4cm}{\textbf{Step Active Minutes 5}} & mean & 0.0419 & 0.123 & 17 & 26.7 \\ 
+    & median & 0 & 0 & 15 & 23\\ 
+    & min & 0 & 0 & 0 & 0 \\ 
+    & max & 11 & 16 & 78 & 213 \\ 
+    \hline
+    \multirow{4}{4cm}{\textbf{Mean Hourly Step}} & mean & 5.88 & 369 & 448 & 756 \\
+    & median & 0 & 366 & 448 & 722 \\
+    & min & 0 & 23 & 111 & 384 \\
+    & max & 401 & 1003 & 777 & 1734 \\
+    \hline
+    \multirow{4}{4cm}{\textbf{Low Active Hours}} & mean & 0.246 & 9.35 & 9.85 & 5.43 \\
+    & median & 0 & 9 & 10 & 5\\
+    & min & 0 & 1 & 2 & 0 \\
+    & max & 8 & 22 & 20 & 13\\
+    \hline
+    \multirow{4}{4cm}{\textbf{Step cadence for periods of moderate and vigorous activity}} & mean & 0.344 & 39 & 63.6 & 77.2 \\
+    & median & 0 & 38.2 & 62.4 & 77.3 \\
+    & min & 41.3 & 125 & 143 & 161 \\
+    & max & 0 & 0 & 0 & 0 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\paragraph{Cluster Metrics}
+Calinski-harabasz index for the steps clusters is 6021.14 .
+
+\begin{table}[H]
+  \caption{Steps clusters Metrics Kmeans}
+  \label{steps_metrics}
+  \centering
+  \begin{tabular}{ c|c|c}
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster name & Cluster Size & Average Silhouette width \\
+    \midrule
+    XS & 1099 & 0.95 \\
+    S & 3942 & 0.34 \\
+    M & 2737 & 0.27 \\
+    L & 1982 & 0.24 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\subsubsection{Hierarchical Clustering with complete linkage}
+\label{appendix:stepsHC}
+The clustering results using Hierarchical clustering with complete linkage on steps data displayed using the first two principal components can be seen in Figure \ref{fig:stepsClustersHC}.
+These clusters were not named and interpreted as the clusters results from Kmeans as the metrics did not shown promising results, hence their default numbering is shown.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[]{steps_HC_results.jpg}
+  \caption{Steps clusters displayed on the first two principle components obtained using Hierarchical clustering}
+  \label{fig:stepsClustersHC}
+\end{figure}
+
+Calinski-harabasz index for the steps clusters is 1518.1865 .
+
+\begin{table}[H]
+  \caption{Steps clusters Metrics HC}
+  \label{steps_metrics}
+  \centering
+  \begin{tabular}{ c|c|c}
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster number & Cluster Size & Average Silhouette width \\
+    \midrule
+    1 & 4996 & 0.12 \\
+    2 & 4056 & 0.11 \\
+    3 & 677 & 0.18 \\
+    4 & 31 & 0.41 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\subsection{Heartrate clusters}
+\label{appendix:hrclusters}
+\subsubsection{K-Means Clustering}
+\label{appendix:hrKmeans}
+\paragraph{Clusters Interpretation}
+Clusters interpretations for the five heartrate clusters can be seen in Table \ref{hr_inter}.
+
+\begin{table}[H]
+  \caption{Heart rate Clusters Interpretation}
+  \label{hr_inter}
+  \centering
+  \begin{tabular}{ c|d }
+    \toprule
+    \cmidrule(r){1-2}
+    \textbf{Cluster name} & \textbf{Cluster Interpretation} \\
+    \midrule
+    S & Small heart rate values and heart rate consistently below 120 \\
+    \midrule
+    SM & Consistent variations around 100 with very few spikes in heart rate values \\
+    \midrule
+    M1 &  Waveform with high variability and periods with spikes above 120 \\
+    \midrule
+    M2 & Consistent variations around 100 with spikes in values above 120 \\
+    \midrule
+    L &  Large values around and above 120 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\paragraph{Clusters Boundaries}
+Clusters boundaries  for the heartrate clusters can be seen in Table \ref{hr_boundaries}.
+
+\begin{table}[H]
+  \caption{Heartrate clusters boundaries - mean, median, min, max}
+  \label{hr_boundaries}
+  \centering
+  \begin{tabular}{p{4cm}*{6}{|p{1cm}}}
+    \toprule
+    \multicolumn{2}{c|}{\backslashbox{Boundaries}{Cluster Name}} & \textbf{S} & \textbf{SM} & \textbf{M1} & \textbf{M2} & \textbf{L} \\
+    \midrule
+    \multirow{4}{4cm}{\textbf{Heartrate Active Minutes 0}} & mean & 4.69 & 14.1 & 60.5 & 50.5 & 134 \\
+    & median & 3 & 13 & 49 & 44 & 111 \\
+    & min & 0 & 0 & 7 &  6 & 20\\
+    & max & 50 & 64 & 726 & 473 & 1439 \\
+    \hline
+    \multirow{4}{4cm}{\textbf{Heartrate Coefficient of Variation}} & mean & 18.1 & 15.9 & 21.4 & 13.6 & 13 \\
+    & median & 18 & 15.9 & 20.9 & 13.4 & 12.7 \\
+    & min & 0 & 7.03 & 14.8 & 6.24 & 10.2 \\
+    & max & 32.9 & 25.7 & 41.6 & 22.1 & 27 \\
+    \hline
+    \multirow{4}{4cm}{\textbf{Number of switches above 100}} & mean & 19.2 & 42.1 & 33.5 & 50.3 & 25.7\\
+    & median & 18 & 41 & 33 & 49 & 26 \\
+    & min & 0 & 12 & 1 & 26 & 0\\
+    & max & 53 & 90 & 68 & 109 & 51 \\
+    \hline
+    \multirow{4}{4cm}{\textbf{Number of switches above 120}} & mean & 1.83 & 6.01 & 16.6 & 17.2 & 33.9 \\
+    & median & 1 & 6 & 15 & 16 & 32 \\
+    & min & 0 & 0 & 1 & 1 & 0\\
+    & max & 8 & 15 & 54 & 52 & 84 \\
+    \hline
+    \multirow{4}{4cm}{\textbf{Median Hourly heartrate}} & mean & 75.6 & 88.8 & 88.7 & 98.6 & 106 \\
+    & median & 76 & 88.8 & 89 & 98.8 & 107\\
+    & min & 48 & 73.1 & 58 & 84.4 & 89.5\\
+    & max & 93.5 & 106 & 118 & 109 & 155 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\paragraph{Cluster Metrics}
+Calinski-harabasz index for the heartrate clusters is 5090.37 .
+
+\begin{table}[H]
+  \caption{Heartrate clusters Metrics Kmeans}
+  \label{hr_metrics}
+  \centering
+  \begin{tabular}{ c|c|c}
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster name & Cluster Size & Average Silhouette width \\
+    \midrule
+    S & 1246 & 0.28 \\
+    SM & 1640 & 0.2 \\
+    M1 & 2308 & 0.21 \\
+    M2 & 2584 & 0.26 \\
+    L & 1982 & 0.33 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\subsubsection{Hierarchical Clustering with complete linkage}
+\label{appendix:hrHC}
+
+The clustering results using Hierarchical clustering with complete linkage for the heart rate data displayed using the first two principal component can be seen in Figure \ref{fig:hrClustersHC}.
+These clusters were not named and interpreted as the clusters results from Kmeans as the metrics did not shown promising results, hence their default numbering is shown.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[]{hr_HC_results.jpg}
+  \caption{Heart rate clusters displayed on the first two principle components obtained using Hierarchical clustering}
+  \label{fig:hrClustersHC}
+\end{figure}
+
+Calinski-Harabasz index for the heart rate clusters is 135.51 .
+
+\begin{table}[H]
+  \caption{Heart rate clusters Metrics HC}
+  \label{hr_metrics}
+  \centering
+  \begin{tabular}{ c|c|c}
+    \toprule
+    \cmidrule(r){1-2}
+    Cluster number & Cluster Size & Average Silhouette width \\
+    \midrule
+    1 & 4221 & -0.07 \\
+    2 & 4901 & -0.06 \\
+    3 & 607 & -0.10 \\
+    4 & 24 & 0.07 \\
+    5 & 7 & -0.06 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\newpage
+\section{Cluster results and metrics for ACT treatments}
+\label{appendix:actclusters}
+
+The tables below show the metrics for the ACT treatment clusters.
+\subsection{K-Means Clustering}
+\label{appendix:actKmeans}
+\subsubsection{K-Means Metrics and Number of Clusters}
+\label{appendix:actKmeansmetrics}
+The table below shows comparison of Average Silhouette width and Calinski-Harabasz index different number of clusters
+
+\begin{table}[H]
+  \caption{Clusters Metrics Kmeans vs Number of Clusters}
+  \label{steps_metrics}
+  \centering
+  \begin{tabular}{ c|c|c}
+    \toprule
+    \cmidrule(r){1-2}
+    Number of Clusters & Calinski-Harabasz index & Average Silhouette width \\
+    \midrule
+    3 & 2292 & 0.17 \\
+    4 & 2208 & 0.18 \\
+    5 & 1996 & 0.15 \\
+    6 & 1830 & 0.15 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\subsubsection{Cluster interpretation}
+\label{appendix:actKmeansinterpretation}
+Clustering with number clusters k=4 has combination of good performance (Average Silhouette width and Calinski-Harabasz index) and positive feedback from clinical team on cluster level granularity and clusters interpretability.
+
+If k is too low, the sub-types are not granular enough, if k is too high it becomes challenging to remember the characteristics of each sub-type.
+
+ Based on these centroids, the four ACT sub-types, which we named alpha, beta, gamma and delta have the following interpretations:
+
+\begin{itemize}
+\item Alpha (Cluster 1) - Treatments with exceptionally big breath counts and biggest treatment adherence.
+\item Beta (Cluster 2) - Breath counts weekly adherence is on a lower side, treatments are short.
+\item Gamma (cluster 3) -  Good breath counts and related breath weekly adherence, the breaths themselves have big standard deviation of duration.
+\item Delta (Cluster 4) - Very good breath counts and related breath weekly adherence,  breaths are mostly short and powerful (with biggest standard deviation for amplitude), shorter treatments.
+\end{itemize}
+
+The clusters capture variability of ACT treatments patient do at home, for example we identified clusters that are characterized by big adherence or breath with higher amplitude.
+
+\begin{figure}[H]
+  \centering
+  \caption{ACT clusters displayed on PCA and UMAP projections obtained using Kmeans clustering}
+  \includegraphics[scale=0.5]{fig_ACT_4_clusters.png}
+  \label{fig:figACT4clusters}
+\end{figure}
+
+\subsection{Hierarchical Clustering}
+\label{appendix:actHC}
+Figure \ref{fig:figACTdent} shows the Dendrogram of ACT data demonstrating major splits for number of clusterts k = 2,3,5 and 6
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.5]{fig_ACT_HC_dentogram.png}
+  \caption{ Hierarchical Clustering Dendrogram for ACT data}
+  \label{fig:figACTdent}
+\end{figure}
+
+The table \ref{tab:act_hc_clusters} shows comparison of Average Silhouette width and Calinski-Harabasz index different number of clusters.
+
+\begin{table}[H]
+  \caption{Clusters Metrics for Hierahical Clustering vs Number of Clusters}
+  \label{tab:act_hc_clusters}
+  \centering
+  \begin{tabular}{ c|c|c}
+    \toprule
+    \cmidrule(r){1-2}
+    Number of Clusters & Calinski-Harabasz index & Average Silhouette width \\
+    \midrule
+    3 & 1793 &  0.13\\
+    4 & 1643 &  0.13\\
+    5 & 1414 &  0.08\\
+    6 & 1289 &  0.06 \\
+    \bottomrule
+    \end{tabular}
+\end{table}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[scale=0.5]{fig_ACT_HC_4_Clusters.png}
+  \caption{ACT clusters displayed on PCA and UMAP projections obtained using Hierarchical clustering}
+  \label{fig:figHC4clusters}
+\end{figure}
+
+\section{ACT Cluster validations via patient trajectories}
+\label{appendix:actpatienttrajectories}
+
+We assessed the validity of the clusters by adding or removing features and checking individual patients’ representation and trajectory in those clusters.
+Figure \ref{fig:ACTclusteringAB} shows the same patient trajectory for clustering with feature selection A and clustering with feature selection B, which includes two additional features: breath amplitude standard deviation, and rolling weekly adherence score for breath count.
+It is notable that once computer gaming was introduced to ACT in both cases A and B the same patient moved from cluster with “expected” breath counts per ACT to a cluster characterized by much higher breath counts.
+This validates that there was a change over time in the patient’s ACT pattern that could be captured.
+The additional breath amplitude standard deviation feature (sdBreathAmplitude) make results more explainable from a clinical point of view, as the original cluster for the patient in Clustering B was characterized by a centroid with the highest sdBreathAmplitude value.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[]{ACTclusteringAB.png}
+  \caption{Patient Trajectory on set A of features (without breath amplitude) and set B of features.}
+  \label{fig:ACTclusteringAB}
+\end{figure}
+
+\section{ACT clustering with different data partitions}
+\label{appendix:actdatapartitions}
+
+We assessed the clustering stability on different data partitions.
+Random sampling 80~\% of the data and clustering using k-means provided very similar results with clustering on the full dataset: figure \ref{fig:ACTclusteringon80} shows that the shape of the clusters, cluster centroids as well as Silhouette Coefficients and Calinski-Harabaz Index (2200 v 1770 respectively) were comparable.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[]{ACTclusteringOn80.png}
+  \caption{K-means clustering results on full dataset (left) and subset of data (right).}
+  \label{fig:ACTclusteringon80}
+\end{figure}
+
 \end{document}
